---
title: "rsmFlow: Foundations to Advanced Response-Surface Workflows"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{rsmFlow: Foundations to Advanced Response-Surface Workflows}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  dpi = 120,
  eval = TRUE
)
```

# rsmFlow: from experimental design to advanced response-surface decisions

**Package:** `rsmFlow`  
**Target version:** `0.2.0-development`  
**Role:** principal starting vignette for students, researchers, instructors, and advanced analysts.  
**Scope:** classical RSM for two or more quantitative factors; two-factor GLM and nonlinear surfaces; optimization, uncertainty, economics, multiresponse analysis, Tier-3 surrogates, and Shiny. Mixture experiments are intentionally outside scope.

> The examples are instructional. Code chunks are not executed automatically in this development snapshot because several advanced engines are optional and full R runtime certification remains a separate release gate. Run the examples locally with the relevant optional packages installed.

## 1. Why this tutorial exists

Response-surface analysis is often taught as a short sequence: fit a quadratic polynomial, draw a contour plot, and report the maximum. That sequence introduces the central geometric idea, but it is not enough for a defensible scientific workflow.

A real experiment can be non-orthogonal, incomplete, constrained by hard factor limits, heteroscedastic, or generated by a response for which Gaussian errors are implausible. A stationary point can lie outside the experimental region. A fitted maximum can occur exactly on the boundary. A biologically interpretable nonlinear model can disagree with a local quadratic approximation. Several responses may conflict. An economic optimum can differ substantially from the biological maximum. A flexible surrogate can improve prediction but produce an unstable decision.

`rsmFlow` therefore organizes response-surface work as an evidence chain:

**Design -> Audit -> Fit -> Diagnose -> Understand geometry -> Optimize within justified bounds -> Quantify uncertainty -> Compare alternatives -> Decide.**

The package is intended to complement established response-surface tools such as `rsm` [@Lenth2009], while integrating design-quality assessment, imperfect/non-orthogonal designs, generalized and nonlinear surfaces, decision-oriented uncertainty, multiresponse analysis, economics, advanced surrogates, and teaching-oriented visualization.

## 2. Learning objectives

After working through this tutorial, the reader should be able to:

1. distinguish natural and coded factor scales;
2. generate CCD, face-centered, inscribed, Box-Behnken, D-optimal, I-optimal, and custom designs;
3. audit rank, aliases, condition number, term correlations, VIF-like inflation, information indices, prediction variance, FDS, VDG, and rotatability;
4. explain why non-orthogonality does not automatically invalidate least squares;
5. fit hierarchical first- and second-order polynomial response surfaces;
6. diagnose residual behavior and lack of fit before optimization;
7. classify stationary points using canonical analysis;
8. distinguish stationary geometry from bounded operational optimization;
9. use steepest ascent/descent and ridge paths;
10. fit two-factor Poisson, binomial, Gamma, inverse-Gaussian, quasi, and negative-binomial response surfaces when appropriate;
11. fit biologically interpretable nonlinear two-factor models;
12. use GA and multistart algorithms for nonlinear starting values without confusing them with the final estimator;
13. quantify uncertainty in optimum coordinates and predicted response;
14. define near-optimal operating zones and optimum stability;
15. distinguish biological and economic optima;
16. use desirability, weighted compromise, distance, epsilon constraints, Pareto fronts, and optional NSGA-II;
17. fit GP, GAM/TPS, RF, and NN surrogates as an explicit expert layer;
18. use expected improvement to recommend the next real experimental run;
19. choose visualizations for geometry, precision, uncertainty, and decision communication;
20. use the optional Shiny interface while preserving a reproducible R-script equivalent.

## 3. The package in one map

| Block | Primary functions | Main scientific question |
|---|---|---|
| Design | `rsm_design()`, `rsm_code()`, `rsm_decode()` | What region and geometry are being studied? |
| Audit | `rsm_design_audit()`, `rsm_orthogonality()` | Is the intended model estimable and numerically stable? |
| Precision | `rsm_prediction_variance()`, `rsm_fds()`, `rsm_vdg()`, `rsm_rotatability()` | Where does the design support precise prediction? |
| Augmentation | `rsm_compare_designs()`, `rsm_augment()` | Which new runs would improve information? |
| Classical RSM | `rsm_fit()`, `rsm_lack_of_fit()`, `rsm_diagnostics()` | Is the polynomial mean surface adequate? |
| Geometry | `rsm_canonical()`, `rsm_steepest()`, `rsm_ridge()` | What shape does the fitted surface imply? |
| Optimization | `rsm_optimize()`, `rsm_robust_optimize()` | What is the best feasible setting? |
| Uncertainty | `rsm_optimum_ci()`, `rsm_optimum_region()`, `rsm_near_optimal()`, `rsm_optimum_stability()` | How uncertain and replaceable is the optimum? |
| GLM-RSM | `rsm_glm_fit()`, `rsm_glm_dispersion()`, `rsm_glm_diagnostics()` | Does the response require a non-Gaussian family? |
| Nonlinear RSM | `rsm_nonlinear_fit()`, `rsm_nonlinear_start_ga()`, `rsm_nonlinear_diagnostics()` | Does the biology suggest saturation, asymptote, or plateau? |
| Multiresponse | `rsm_desirability()`, `rsm_multiopt()`, `rsm_pareto()` | How are conflicting responses reconciled? |
| Economics | `rsm_economic_optimum()` | Does input cost change the recommendation? |
| Tier 3 | `rsm_surrogate()`, `rsm_surrogate_optimize()`, `rsm_bayes_opt()` | Is a more flexible predictor justified? |
| Comparison | `rsm_cv()`, `rsm_compare_engines()`, `rsm_compare_optima()` | Do alternative models lead to the same decision? |
| Graphics | `rsm_plot()`, `rsm_slices()`, `rsm_profile()`, `rsm_perturbation()` | Which view answers the scientific question? |
| Interface | `run_rsm_app()` | Can the same workflow be taught interactively? |

# Part I. Design before modeling

## 4. Load the bundled agronomic example

```{r}
library(rsmFlow)

path <- system.file(
  "extdata",
  "agronomy_rsm_nonorthogonal.csv",
  package = "rsmFlow"
)

dat <- read.csv(path)
str(dat)
head(dat)
summary(dat)
```

This dataset is useful because it is not an idealized perfectly orthogonal textbook design. It supports the package's central teaching principle: **a design should be judged by the information it provides for the intended model and decision, not by a single orthogonality label.**

Before fitting, record a table like this:

| Item | What to record |
|---|---|
| Response | name, unit, meaningful range |
| Factors | names and natural units |
| Bounds | actual experimental minimum and maximum |
| Coding | center and scale, if used |
| Replication | repeated factor combinations / center points |
| Blocking | block or batch structure |
| Constraints | infeasible combinations or hard process limits |
| Design provenance | planned CCD/BBD/optimal design or custom/imperfect design |

## 5. Code and decode factor scales

If a natural factor is centered at \(X_0\) with scale \(\Delta X\),

\[
x = \frac{X-X_0}{\Delta X}.
\]

```{r}
coded <- rsm_code(
  dat,
  factors = c("N", "K", "Irrigation")
)

head(coded)
coding <- attr(coded, "rsmFlow_coding")
coding
```

Decode:

```{r}
restored <- rsm_decode(
  coded,
  coding
)
head(restored)
```

**Interpretation.** Coding changes coordinates, not the biological identity of a factor. Report recommendations in natural units even if the internal geometry is studied on coded scales.

## 6. Generate a central composite design

```{r}
ccd <- rsm_design(
  factors = c("N", "P"),
  type = "ccd",
  n0 = 5,
  alpha = "rotatable",
  seed = 260820
)
ccd
```

## 7. Face-centered and inscribed CCDs

```{r}
ccf <- rsm_design(
  factors = c("N", "P", "Water"),
  type = "ccd_face",
  n0 = 4,
  seed = 260820
)

cci <- rsm_design(
  factors = c("N", "P", "Water"),
  type = "ccd_inscribed",
  n0 = 4,
  seed = 260820
)
```

Use face-centered or inscribed variants when axial points from a circumscribed CCD would exceed safe or feasible factor limits.

## 8. Box-Behnken design

```{r}
bbd <- rsm_design(
  factors = c("N", "P", "K"),
  type = "bbd",
  n0 = 5,
  seed = 260820
)
```

Box-Behnken designs avoid the full corners of a three-level cube, which can be valuable when simultaneous extremes are difficult or unsafe.

## 9. D-optimal design from a candidate region

```{r}
candidates <- expand.grid(
  N = seq(-1, 1, length.out = 9),
  P = seq(-1, 1, length.out = 9),
  K = seq(-1, 1, length.out = 7)
)

# Example practical exclusion.
candidates <- subset(
  candidates,
  !(N > .75 & P > .75 & K > .75)
)

d_opt <- rsm_design(
  factors = c("N", "P", "K"),
  type = "d_optimal",
  runs = 24,
  candidates = candidates,
  seed = 260820
)
```

## 10. I-optimal design

```{r}
i_opt <- rsm_design(
  factors = c("N", "P", "K"),
  type = "i_optimal",
  runs = 24,
  candidates = candidates,
  seed = 260820
)
```

**Interpretation.** D-oriented criteria emphasize parameter information. I-oriented criteria emphasize average prediction variance. The preferred criterion should match the scientific objective and the exact implementation/backend should be validated before formal optimal-design claims.

## 11. Fit the intended second-order model

```{r}
fit <- rsm_fit(
  dat,
  response = "Yield",
  factors = c("N", "K", "Irrigation"),
  order = 2
)

fit
summary(fit)
```

The hierarchical second-order surface is

\[
\hat y(\mathbf x)
= \beta_0
+ \sum_i \beta_i x_i
+ \sum_i \beta_{ii}x_i^2
+ \sum_{i<j}\beta_{ij}x_ix_j.
\]

Hierarchy matters because quadratic and interaction terms are interpreted in the presence of their corresponding lower-order terms.

## 12. Audit the model matrix

```{r}
audit <- rsm_design_audit(fit)
audit
```

Inspect the principal components:

```{r}
audit$rank
audit$estimable
audit$exact_alias
audit$condition_number_scaled
audit$vif
audit$orthogonality$correlation
audit$information_indices
```

## 13. Why non-orthogonality is not automatic invalidity

If \(X\) is full rank, ordinary least squares remains

\[
\hat{\boldsymbol\beta}
=(X^\top X)^{-1}X^\top y.
\]

The practical problem is loss of precision and interpretability as columns of \(X\) become correlated. Response-surface design work using generalized variance-inflation concepts treats this as an information penalty rather than a binary invalid/valid distinction [@ODriscoll2015].

A strong report should state:

1. whether all planned terms are estimable;
2. which model directions are highly correlated;
3. whether coefficient uncertainty is inflated;
4. where prediction variance becomes large;
5. whether the intended decision region is well supported.

## 14. Prediction variance

For regressor vector \(f(\mathbf x)\), scaled prediction variance is proportional to

\[
f(\mathbf x)^\top (X^\top X)^{-1} f(\mathbf x).
\]

```{r}
pv <- rsm_prediction_variance(
  fit,
  n = 35,
  scaled = TRUE
)
head(pv)
```

```{r}
rsm_plot(
  fit,
  type = "prediction_variance"
)
```

## 15. FDS and VDG

```{r}
fds <- rsm_fds(
  fit,
  n = 20000,
  seed = 260820
)
plot(fds)
```

```{r}
vdg <- rsm_vdg(
  fit,
  radii = seq(0, 1, length.out = 21),
  points = 1500,
  seed = 260820
)
plot(vdg)
```

FDS answers how much of the region achieves a given prediction-variance level. VDG shows how variance behaves over shells moving away from the center.

## 16. Rotatability

```{r}
rot <- rsm_rotatability(
  fit,
  radii = c(.25, .5, .75, 1),
  points = 1500,
  seed = 260820
)
rot
```

Rotatability is useful when equal-distance precision is important, but it is not universally superior to a design that intentionally places more precision in a practically important subregion.

## 17. Augment the design without altering existing observations

```{r}
add_D <- rsm_augment(
  fit,
  n_add = 4,
  objective = "D",
  seed = 260820
)
```

Prediction-oriented:

```{r}
add_I <- rsm_augment(
  fit,
  n_add = 4,
  objective = "I",
  seed = 260820
)
```

Worst-variance oriented:

```{r}
add_G <- rsm_augment(
  fit,
  n_add = 4,
  objective = "G",
  seed = 260820
)
```

**Interpretation.** Augmentation proposes new experimental runs. It should never “correct” observed responses or remove inconvenient runs solely to improve orthogonality.

# Part II. Fit and diagnose classical surfaces

## 18. First-order versus second-order models

```{r}
fit_fo <- rsm_fit(
  dat,
  "Yield",
  c("N", "K", "Irrigation"),
  order = 1
)

fit_so <- rsm_fit(
  dat,
  "Yield",
  c("N", "K", "Irrigation"),
  order = 2
)
```

A first-order model is particularly valuable during sequential exploration when curvature is negligible. A second-order model adds local curvature and interaction geometry.

## 19. Weighted and robust sensitivity fits

```{r}
fit_wls <- rsm_fit(
  dat,
  "Yield",
  c("N", "K", "Irrigation"),
  order = 2,
  estimator = "wls_power"
)
```

```{r}
fit_robust <- rsm_fit(
  dat,
  "Yield",
  c("N", "K", "Irrigation"),
  order = 2,
  estimator = "robust"
)
```

Use them to address a diagnosed variance or influence problem, not to search for a preferred significance pattern.

## 20. Residual diagnostics

```{r}
diag <- rsm_diagnostics(fit_so)
diag
```

```{r}
rsm_plot(fit_so, "residuals")
rsm_plot(fit_so, "qq")
rsm_plot(fit_so, "leverage")
rsm_plot(fit_so, "cooks")
```

Read these for unmodeled curvature, heterogeneous variance, unusual tails, influential runs, and design regions with disproportionate leverage.

## 21. Lack of fit

```{r}
lof <- rsm_lack_of_fit(fit_so)
lof
```

When replicated factor combinations exist, lack of fit contrasts model residual variation with pure error. A non-significant result does not prove that the model is true; a significant result is evidence that the fitted mean structure is inadequate relative to replicate variation.

# Part III. Geometry

## 22. Canonical analysis

```{r}
ca <- rsm_canonical(fit_so)
ca
ca$stationary_point
ca$eigenvalues
ca$nature
```

For

\[
\hat y(\mathbf x)
= \hat\beta_0 + \mathbf b^\top\mathbf x + \mathbf x^\top B\mathbf x,
\]

the stationary point is

\[
\mathbf x_s=-\frac12B^{-1}\mathbf b
\]

when \(B\) is nonsingular.

| Eigenvalue pattern | Canonical interpretation |
|---|---|
| all negative | local maximum |
| all positive | local minimum |
| mixed signs | saddle |
| near-zero direction | ridge / weak curvature |

## 23. Three synthetic geometry examples

### Maximum

```{r}
set.seed(1)
dmax <- expand.grid(x1 = seq(-1,1,length.out=7), x2 = seq(-1,1,length.out=7))
dmax$y <- 10 - 2*dmax$x1^2 - 3*dmax$x2^2 + rnorm(nrow(dmax),0,.08)
rsm_canonical(rsm_fit(dmax, "y", c("x1","x2")))
```

### Minimum

```{r}
set.seed(2)
dmin <- expand.grid(x1 = seq(-1,1,length.out=7), x2 = seq(-1,1,length.out=7))
dmin$y <- 4 + 2*dmin$x1^2 + 1.5*dmin$x2^2 + rnorm(nrow(dmin),0,.08)
rsm_canonical(rsm_fit(dmin, "y", c("x1","x2")))
```

### Saddle

```{r}
set.seed(3)
ds <- expand.grid(x1 = seq(-1,1,length.out=7), x2 = seq(-1,1,length.out=7))
ds$y <- 8 + 2*ds$x1^2 - 2*ds$x2^2 + rnorm(nrow(ds),0,.08)
rsm_canonical(rsm_fit(ds, "y", c("x1","x2")))
```

A saddle is not an optimum.

## 24. Steepest ascent and descent

```{r}
up <- rsm_steepest(
  fit_fo,
  direction = "ascent",
  step = .1,
  n_steps = 12
)
up
```

```{r}
down <- rsm_steepest(
  fit_fo,
  direction = "descent",
  step = .1,
  n_steps = 12
)
```

These paths recommend directions for new experimentation. They are not observed response sequences until the proposed runs are physically conducted.

## 25. Ridge analysis

```{r}
ridge <- rsm_ridge(
  fit_so,
  radii = seq(.1, 1, by = .1),
  goal = "max",
  n_starts = 30,
  seed = 260820
)
ridge
```

Ridge analysis is especially useful when the stationary point lies outside the design region or curvature is weak.

# Part IV. Bounded optimization and uncertainty

## 26. Bounded maximum

```{r}
opt <- rsm_optimize(
  fit_so,
  goal = "max",
  method = "hybrid",
  seed = 260820
)
opt
```

Compare methods:

```{r}
rsm_optimize(fit_so, "max", method = "analytic")
rsm_optimize(fit_so, "max", method = "grid", grid_n = 41)
rsm_optimize(fit_so, "max", method = "L-BFGS-B")
rsm_optimize(fit_so, "max", method = "GA", seed = 260820)
rsm_optimize(fit_so, "max", method = "all", seed = 260820)
```

Agreement on a smooth problem supports numerical reliability. Disagreement should trigger investigation of flatness, multimodality, bounds, or optimizer settings.

## 27. Target optimization

```{r}
target <- rsm_optimize(
  fit_so,
  goal = "target",
  target = 8.5,
  method = "hybrid",
  seed = 260820
)
```

Target optimization is appropriate for process variables such as pH, moisture, hardness, or concentration when “more” is not always better.

## 28. Bootstrap optimum uncertainty

```{r, eval=FALSE}
unc <- rsm_optimum_ci(
  fit_so,
  optimum = opt,
  B = 1999,
  conf = .95,
  method = "residual",
  optimizer = "L-BFGS-B",
  seed = 260820
)
unc$intervals
unc$failures
```

Formal confidence sets for optimum factor levels are a recognized inference problem [@Wan2016]. The package's bootstrap layer is a transparent computational approximation and should be described as such.

## 29. Joint region and optimum cloud

```{r, eval=FALSE}
region <- rsm_optimum_region(
  unc,
  conf = .95,
  n = 300
)
plot(region)
```

```{r, eval=FALSE}
rsm_plot_optimum_uncertainty(unc)
```

A correlated/elongated cloud cannot be summarized fully by two separate coordinate intervals.

## 30. Near-optimal operating region

```{r}
near <- rsm_near_optimal(
  fit_so,
  optimum = opt,
  tolerance = .05,
  n = 45,
  seed = 260820
)
head(near)
```

A near-optimal region can reveal practical alternatives that sacrifice very little fitted response while reducing cost or operational difficulty.

## 31. Optimum stability

```{r, eval=FALSE}
stability <- rsm_optimum_stability(
  unc,
  fit_so,
  tolerance = .10
)
stability
```

## 32. Uncertainty-aware optimization

```{r}
conservative <- rsm_robust_optimize(
  fit_so,
  goal = "max",
  lambda = 1.96,
  method = "L-BFGS-B",
  seed = 260820
)
```

This trades fitted mean for prediction support. It is a decision criterion, not an exact confidence guarantee.

# Part V. GLM response surfaces

## 33. Why GLM-RSM?

A second-order predictor can be retained while the response family changes:

\[
g(\mu)=\beta_0+\beta_1x_1+\beta_2x_2+\beta_{11}x_1^2+\beta_{22}x_2^2+\beta_{12}x_1x_2.
\]

Use the response family to reflect the data-generating process.

## 34. Poisson example: insect counts

```{r}
set.seed(260820)
insects <- expand.grid(
  Dose = seq(0,1,length.out=8),
  Time = seq(0,1,length.out=7)
)

eta <- 3.4 - 1.3*insects$Dose + 1.1*insects$Time -
  .9*insects$Time^2 + .45*insects$Dose*insects$Time

insects$Count <- rpois(nrow(insects), exp(eta))

pois <- rsm_glm_fit(
  insects,
  "Count",
  c("Dose","Time"),
  family = poisson()
)

rsm_glm_dispersion(pois)
rsm_glm_diagnostics(pois)
```

## 35. Binomial example with trial weights

```{r}
set.seed(260821)
germ <- expand.grid(
  Temperature = seq(18,34,length.out=9),
  Salinity = seq(0,8,length.out=7)
)

lp <- 2.4 - .045*(germ$Temperature-26)^2 - .32*germ$Salinity
p <- plogis(lp)
germ$Trials <- 50
germ$Proportion <- rbinom(nrow(germ), germ$Trials, p) / germ$Trials

bin <- rsm_glm_fit(
  germ,
  "Proportion",
  c("Temperature","Salinity"),
  family = binomial(),
  weights = germ$Trials
)
```

## 36. Gamma example

```{r}
set.seed(260822)
gd <- expand.grid(
  Temperature = seq(20,38,length.out=8),
  Moisture = seq(.3,.9,length.out=7)
)

mu <- exp(2 + .035*(gd$Temperature-28) -
  .0025*(gd$Temperature-28)^2 + .9*gd$Moisture)

gd$Respiration <- rgamma(nrow(gd), shape=30, scale=mu/30)

gamma_fit <- rsm_glm_fit(
  gd,
  "Respiration",
  c("Temperature","Moisture"),
  family = Gamma(link="log")
)
```

## 37. Negative binomial and offsets

```{r}
if (requireNamespace("MASS", quietly = TRUE)) {
  nb <- rsm_glm_fit(
    insects,
    "Count",
    c("Dose","Time"),
    family = "negative_binomial"
  )
}
```

```{r}
insects$Exposure <- sample(c(5,10,20), nrow(insects), replace=TRUE)
rate_fit <- rsm_glm_fit(
  insects,
  "Count",
  c("Dose","Time"),
  family = poisson(),
  offset = log(insects$Exposure),
  offset_reference = log(10)
)
```

**Interpretation.** Optimize and plot GLM surfaces on the scientifically meaningful response scale unless the link scale itself is the stated objective.

## 38. GLM canonical and optimum uncertainty

```{r, eval=FALSE}
rsm_glm_canonical(pois)

pois_opt <- rsm_optimize(
  pois,
  goal = "min",
  scale = "response"
)

pois_unc <- rsm_optimum_ci(
  pois,
  optimum = pois_opt,
  B = 999,
  seed = 260820
)
```

For quasi families, avoid likelihood-based rankings that presume an ordinary likelihood.

# Part VI. Nonlinear biological response surfaces

## 39. Inspect the nonlinear catalog

```{r}
rsm_nonlinear_models()
```

The package includes Mitscherlich, Gompertz, Michaelis-Menten, asymptotic, Hoerl, linear plateau, quadratic plateau, and custom formulas.

## 40. Mitscherlich with GA-assisted starts

```{r}
set.seed(260820)
nl_dat <- expand.grid(
  N = seq(0,160,length.out=9),
  P = seq(0,90,length.out=8)
)

mu <- 11*(1-exp(-.025*nl_dat$N))*(1-exp(-.04*nl_dat$P))
nl_dat$Yield <- mu + rnorm(nrow(nl_dat),0,.12)

fit_nl <- rsm_nonlinear_fit(
  nl_dat,
  "Yield",
  c("N","P"),
  model = "mitscherlich2_product",
  start = "ga",
  seed = 260820
)

rsm_nonlinear_diagnostics(fit_nl)
```

## 41. Michaelis-Menten with multistart

```{r}
set.seed(260821)
mm <- expand.grid(
  N = seq(5,150,length.out=9),
  P = seq(5,100,length.out=8)
)

mm$Yield <- 12*mm$N/(45+mm$N)*mm$P/(30+mm$P) +
  rnorm(nrow(mm),0,.10)

fit_mm <- rsm_nonlinear_fit(
  mm,
  "Yield",
  c("N","P"),
  model = "michaelis_menten2",
  start = "multistart",
  n_start = 1000,
  seed = 260820
)
```

## 42. Custom nonlinear model

```{r, eval=FALSE}
custom_formula <- Yield ~
  A*(1-exp(-kN*N))*exp(-kW*(Water-Wopt)^2)

custom_fit <- rsm_nonlinear_fit(
  custom_data,
  "Yield",
  c("N","Water"),
  model = "custom",
  formula = custom_formula,
  start = "ga",
  start_bounds = list(
    A = c(5,20),
    kN = c(.001,.1),
    kW = c(.001,.2),
    Wopt = c(.3,1.0)
  ),
  seed = 260820
)
```

GA searches for useful starting values. Final estimation remains `nls`, `nlsLM`, or `gnls`.

## 43. Parameter identifiability

```{r}
nld <- rsm_nonlinear_diagnostics(fit_nl)
nld$jacobian_condition
nld$parameter_correlation
nld$converged
```

A low residual error does not guarantee that nonlinear parameters are well identified.

## 44. Generalized nonlinear least squares

```{r, eval=FALSE}
if (requireNamespace("nlme", quietly=TRUE)) {
  gnls_fit <- rsm_nonlinear_fit(
    nl_dat,
    "Yield",
    c("N","P"),
    model = "mitscherlich2_product",
    start = "ga",
    engine = "gnls",
    variance = "power",
    seed = 260820
  )
}
```

Use this only when variance/correlation structure is scientifically supported.

# Part VII. Economics and multiple responses

## 45. Biological versus economic optimum

```{r}
eco <- rsm_economic_optimum(
  fit_so,
  response_price = 1200,
  factor_cost = c(
    N = 5.5,
    K = 4.0,
    Irrigation = 900
  ),
  fixed_cost = 350,
  seed = 260820
)
eco
```

The objective is

\[
\Pi(\mathbf x)
=p_y\hat y(\mathbf x)-\sum_i c_i(x_i-x_{i,0})-C_0.
\]

Costs are interpreted relative to a baseline so coded negative factor values do not generate nonsensical negative input costs.

## 46. Price/cost sensitivity

```{r}
price_grid <- c(600,900,1200,1500)

price_sensitivity <- lapply(price_grid, function(p) {
  rsm_economic_optimum(
    fit_so,
    response_price = p,
    factor_cost = c(N=5.5,K=4.0,Irrigation=900),
    fixed_cost = 350,
    seed = 260820
  )
})
```

Report sensitivity when economic assumptions materially change the recommended region.

## 47. Multiresponse data and fitted surfaces

```{r}
set.seed(260820)
multi <- expand.grid(
  N = seq(40,160,length.out=9),
  Water = seq(.5,1.3,length.out=8)
)

multi$Yield <- 6 + .06*multi$N + 5*multi$Water -
  .00022*multi$N^2 - 2*multi$Water^2 + .006*multi$N*multi$Water +
  rnorm(nrow(multi),0,.12)

multi$Quality <- 78 + .045*multi$N + 4*multi$Water -
  .00018*multi$N^2 - 2.5*(multi$Water-.9)^2 +
  rnorm(nrow(multi),0,.20)

multi$Disease <- 20 - .04*multi$N - 8*multi$Water +
  .00020*multi$N^2 + 5*multi$Water^2 +
  rnorm(nrow(multi),0,.25)

fit_y <- rsm_fit(multi,"Yield",c("N","Water"))
fit_q <- rsm_fit(multi,"Quality",c("N","Water"))
fit_d <- rsm_fit(multi,"Disease",c("N","Water"))
```

Each response should be diagnosed independently before it enters a joint decision.

## 48. Desirability

```{r}
res_des <- rsm_multiopt(
  list(fit_y,fit_q,fit_d),
  goals = c("max","target","min"),
  limits = list(
    list(low=9, high=13),
    list(low=78, target=82, high=88),
    list(low=5, high=20)
  ),
  method = "desirability",
  importance = c(1.5,1,1.2),
  seed = 260820
)
```

The desirability framework follows @DerringerSuich1980. Bounds, targets, and importance values are decision preferences and must be reported.

## 49. Weighted, distance, and epsilon methods

```{r}
res_weighted <- rsm_multiopt(
  list(fit_y,fit_d),
  goals = c("max","min"),
  limits = list(
    list(low=9,high=13),
    list(low=5,high=20)
  ),
  method = "weighted",
  importance = c(2,1),
  seed = 260820
)
```

```{r}
res_epsilon <- rsm_multiopt(
  list(fit_y,fit_d),
  goals = c("max","min"),
  limits = list(
    list(low=9,high=13),
    list(low=5,high=20)
  ),
  method = "epsilon",
  primary = 1,
  epsilon = c(NA,10),
  seed = 260820
)
```

## 50. Pareto and optional NSGA-II

```{r}
pareto <- rsm_pareto(
  list(fit_y,fit_d),
  goals = c("max","min"),
  n = 20000,
  engine = "sample",
  seed = 260820
)

rsm_plot_pareto(pareto)
```

```{r}
if (requireNamespace("mco", quietly=TRUE)) {
  pareto_nsga <- rsm_pareto(
    list(fit_y,fit_d),
    goals = c("max","min"),
    engine = "NSGA2",
    seed = 260820
  )
}
```

A Pareto front is a set of efficient trade-offs, not one universal optimum.

# Part VIII. Tier-3 surrogates and sequential optimization

## 51. Why Tier 3 is explicit

A quadratic response surface remains the default when it is adequate because coefficients and canonical geometry are directly interpretable. Advanced surrogates are a sensitivity/prediction layer for problems where simpler models remain inadequate.

## 52. GP, GAM/TPS, RF, NN

```{r}
if (requireNamespace("DiceKriging", quietly=TRUE)) {
  gp <- rsm_surrogate(
    dat,
    "Yield",
    c("N","K","Irrigation"),
    method = "gp",
    seed = 260820
  )
}
```

```{r}
if (requireNamespace("mgcv", quietly=TRUE)) {
  gam <- rsm_surrogate(dat,"Yield",c("N","K","Irrigation"),method="gam")
  tps <- rsm_surrogate(dat,"Yield",c("N","K","Irrigation"),method="tps")
}
```

```{r}
if (requireNamespace("ranger", quietly=TRUE)) {
  rf_fit <- rsm_surrogate(dat,"Yield",c("N","K","Irrigation"),method="rf")
}

if (requireNamespace("nnet", quietly=TRUE)) {
  nn <- rsm_surrogate(dat,"Yield",c("N","K","Irrigation"),method="nn")
}
```

## 53. Structured cross-validation

```{r}
cv <- rsm_cv(
  fit_so,
  folds = 5,
  repeats = 10,
  seed = 260820
)
cv
```

When experimental blocks should remain intact:

```{r, eval=FALSE}
cv_block <- rsm_cv(
  blocked_fit,
  folds = 5,
  repeats = 5,
  group = "Block",
  seed = 260820
)
```

Random row-wise cross-validation is not universally appropriate for designed experiments.

## 54. Compare engines and decisions separately

```{r, eval=FALSE}
rsm_compare_engines(
  fit_so,
  fit_nl,
  gp
)
```

```{r, eval=FALSE}
opt_classical <- rsm_optimize(fit_so, seed=260820)
opt_nonlinear <- rsm_optimize(fit_nl, seed=260820)
opt_gp <- rsm_surrogate_optimize(gp, seed=260820)

rsm_compare_optima(
  opt_classical,
  opt_nonlinear,
  opt_gp
)
```

Similar predictive error can produce different recommendations. Conversely, very different models can lead to essentially the same operating region.

## 55. Bayesian optimization

```{r, eval=FALSE}
if (requireNamespace("DiceKriging", quietly=TRUE)) {
  next_run <- rsm_bayes_opt(
    dat,
    "Yield",
    c("N","K","Irrigation"),
    goal = "max",
    n_candidates = 20000,
    xi = .01,
    seed = 260820
  )
  next_run$next_point
}
```

Expected Improvement follows the efficient global optimization framework [@JonesSchonlauWelch1998]. The proposed point is **not an observation**. Conduct the experiment, append the measured response, refit, and only then request another point.

# Part IX. Visualization and teaching

## 56. Contours, heat maps, and surfaces

```{r}
rsm_plot(fit_so, type="contour", x="N", y="K")
rsm_plot(fit_so, type="heatmap", x="N", y="K")
```

```{r}
if (requireNamespace("plotly", quietly=TRUE)) {
  rsm_plot(
    fit_so,
    type="surface",
    x="N",
    y="K",
    interactive=TRUE
  )
}
```

## 57. Slices for three or more factors

```{r}
slices <- rsm_slices(
  fit_so,
  x="N",
  y="K",
  slice_factor="Irrigation",
  values=quantile(dat$Irrigation,c(.2,.5,.8)),
  n=60
)
plot(slices)
```

A slice is conditional. State the fixed value of the third factor in every panel/caption.

## 58. Profiles and perturbation

```{r}
prof <- rsm_profile(
  fit_so,
  factor="N",
  at=c(K=median(dat$K),Irrigation=median(dat$Irrigation)),
  n=150
)
plot(prof)
```

```{r}
pert <- rsm_perturbation(fit_so)
plot(pert)
```

Profiles show conditional response curves. Perturbation curves show local sensitivity around a common reference point; they are not global variable importance.

## 59. Overlaid constraints and desirability surfaces

```{r}
over <- rsm_overlaid_contour(
  list(fit_y,fit_q,fit_d),
  constraints=list(
    list(goal="max",low=11),
    list(goal="max",low=80),
    list(goal="min",high=10)
  )
)
plot(over)
```

```{r}
dsurf <- rsm_desirability_surface(
  list(fit_y,fit_q,fit_d),
  goals=c("max","target","min"),
  limits=list(
    list(low=9,high=13),
    list(low=78,target=82,high=88),
    list(low=5,high=20)
  )
)
plot(dsurf)
```

## 60. Shiny as a thin teaching layer

```{r, eval=FALSE}
if (requireNamespace("shiny", quietly=TRUE)) {
  run_rsm_app()
}
```

The interface should be used to expose the same tested API, not to create a second hidden implementation.

A useful teaching pattern is:

| GUI action | Script equivalent |
|---|---|
| Fit classical RSM | `rsm_fit()` |
| Audit design | `rsm_design_audit()` |
| Diagnose | `rsm_diagnostics()` |
| Canonical analysis | `rsm_canonical()` |
| Optimize | `rsm_optimize()` |
| Plot contour | `rsm_plot()` |

## 61. Use `rsm_explain()` after numerical output

```{r}
cat(rsm_explain(fit_so), sep="\n")
```

Instructional explanation should reinforce the numerical object, not replace it.

# Part X. Three complete workflows

## 62. Beginner workflow: classical RSM

```{r, eval=FALSE}
library(rsmFlow)

dat <- read.csv(
  system.file("extdata","agronomy_rsm_nonorthogonal.csv",package="rsmFlow")
)

fit <- rsm_fit(dat,"Yield",c("N","K","Irrigation"),order=2)
audit <- rsm_design_audit(fit)
diag <- rsm_diagnostics(fit)
ca <- rsm_canonical(fit)
opt <- rsm_optimize(fit,"max",method="hybrid",seed=260820)
near <- rsm_near_optimal(fit,opt,tolerance=.05,seed=260820)

print(audit)
print(diag)
print(ca)
print(opt)
head(near)
rsm_plot(fit,"contour",optimum=opt)
```

**Interpretation sequence:** design information -> adequacy -> geometry -> bounded optimum -> boundary status -> near-optimal alternatives.

## 63. Intermediate workflow: nonlinear biology

```{r, eval=FALSE}
fit_nl <- rsm_nonlinear_fit(
  nl_dat,
  "Yield",
  c("N","P"),
  model="mitscherlich2_product",
  start="ga",
  seed=260820
)

nl_diag <- rsm_nonlinear_diagnostics(fit_nl)
nl_opt <- rsm_optimize(fit_nl,method="hybrid",seed=260820)
nl_unc <- rsm_optimum_ci(fit_nl,nl_opt,B=999,seed=260820)

eco_nl <- rsm_economic_optimum(
  fit_nl,
  response_price=1200,
  factor_cost=c(N=5.5,P=4.0),
  fixed_cost=250,
  seed=260820
)
```

**Interpretation sequence:** biological model justification -> start search -> convergence/identifiability -> bounded response optimum -> uncertainty -> economic decision.

## 64. Advanced workflow: multiresponse plus Tier 3

```{r, eval=FALSE}
res <- rsm_multiopt(
  list(fit_y,fit_q,fit_d),
  goals=c("max","target","min"),
  limits=list(
    list(low=9,high=13),
    list(low=78,target=82,high=88),
    list(low=5,high=20)
  ),
  method="desirability",
  importance=c(1.5,1,1.2),
  seed=260820
)

pareto <- rsm_pareto(
  list(fit_y,fit_d),
  goals=c("max","min"),
  n=20000,
  seed=260820
)

if (requireNamespace("DiceKriging", quietly=TRUE)) {
  gp <- rsm_surrogate(multi,"Yield",c("N","Water"),method="gp",seed=260820)
  next_run <- rsm_bayes_opt(multi,"Yield",c("N","Water"),seed=260820)
}
```

**Interpretation sequence:** validate each response -> expose preferences -> visualize Pareto alternatives -> validate advanced predictor -> treat next-run recommendation as a proposal for a real new experiment.

# Part XI. Common mistakes

## 65. “Non-orthogonal means invalid”

Use rank, VIF-like inflation, prediction variance, FDS, and VDG. Non-orthogonality is a measurable information problem, not a binary rejection rule.

## 66. Optimizing before diagnostics

A successful optimizer can maximize a scientifically inadequate model.

## 67. Calling every stationary point an optimum

A saddle is not an optimum; an outside stationary point is not an experimentally established recommendation.

## 68. Extrapolating without disclosure

Default operational bounds should reflect the region actually studied. If extrapolation is scientifically justified, label it clearly.

## 69. Treating every response as Gaussian

Use GLM or nonlinear models when the response process justifies them.

## 70. Confusing GA with inference

For nonlinear fitting, GA searches starting values; the final model is estimated by the nonlinear regression engine.

## 71. Reporting only one optimum coordinate

Use optimum uncertainty, joint regions, near-optimal zones, and stability.

## 72. Treating multiresponse preferences as statistical truth

Weights, targets, and desirability limits are explicit decision assumptions.

## 73. Choosing Tier 3 from training error

Use structured validation, model adequacy, interpretability, and decision stability.

## 74. Treating a Bayesian-optimization point as measured data

The point must be physically tested.

# Part XII. Reporting checklist

## 75. Minimum information for a defensible analysis

- [ ] response and factor units;
- [ ] experimental natural bounds;
- [ ] coding transformation;
- [ ] design provenance;
- [ ] planned model order/family;
- [ ] rank and estimability;
- [ ] important conditioning/correlation findings;
- [ ] prediction-variance behavior;
- [ ] residual/model-family diagnostics;
- [ ] lack of fit where pure error exists;
- [ ] canonical classification for SO models;
- [ ] stationary point inside/outside status;
- [ ] optimization bounds and method;
- [ ] boundary status;
- [ ] uncertainty method, B, confidence level, seed, failures;
- [ ] near-optimal tolerance;
- [ ] economic assumptions if used;
- [ ] multiresponse targets/weights/constraints;
- [ ] validation strategy for Tier 3;
- [ ] acquisition method/candidate region for Bayesian optimization;
- [ ] optional backend versions;
- [ ] limits on extrapolation.

## 76. Recommended next vignettes

The detailed curriculum is described in `VIGNETTE_ORGANIZATION.md`.

1. `01-design-audit-and-augmentation.Rmd` - design information and repair;
2. `02-classical-rsm-geometry-diagnostics.Rmd` - polynomial hierarchy, diagnostics, canonical geometry;
3. `03-glm-and-nonlinear-surfaces.Rmd` - non-Gaussian and biological surfaces;
4. `04-optimization-uncertainty-economics.Rmd` - decision, bootstrap, operating zones, economics;
5. `05-multiresponse-and-tier3.Rmd` - preferences, Pareto, surrogates, Bayesian optimization;
6. `06-visualization-shiny-and-teaching.Rmd` - graphics and interactive pedagogy;
7. `07-validation-reproducibility-and-method-selection.Rmd` - simulations, CV, method comparison, release evidence.

# Final perspective

The 53 exported functions in `rsmFlow` are not 53 unrelated commands. They express one progressive response-surface grammar:

**define the region -> generate/import the design -> audit information -> fit the simplest scientifically adequate response model -> diagnose -> understand geometry -> optimize inside justified bounds -> propagate uncertainty -> identify practical alternatives -> incorporate economics and conflicting responses -> escalate model flexibility only when justified -> propose new experiments without fabricating observations -> communicate the complete evidence chain.**

For a beginner, the central lesson is that a contour plot is the middle of the analysis, not the beginning or the end.

For an advanced user, the central lesson is that additional flexibility is useful only when it changes prediction, uncertainty, or scientific decision in a validated and interpretable way.
