---
title: "Non-orthogonal and imperfect response-surface designs"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Non-orthogonal and imperfect response-surface designs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
---

A non-orthogonal design does not automatically invalidate least-squares response-surface fitting. If the model matrix \(X\) has full column rank, the OLS estimator remains

\[
\widehat{\boldsymbol\beta}=(X^\top X)^{-1}X^\top y.
\]

The practical problem is loss of precision and interpretability as columns of the model matrix become strongly correlated. This is why `rsm_design_audit()` evaluates rank, condition number, VIF, model-term correlations, information eigenvalues, prediction variance and Fraction-of-Design-Space summaries. The GVIF literature explicitly treats non-orthogonality as a variance-inflation penalty in response-surface designs [@ODriscoll2015].

For a point \(\mathbf x\), the scaled prediction variance is proportional to

\[
f(\mathbf x)^\top(X^\top X)^{-1}f(\mathbf x).
\]

The package therefore emphasizes prediction-variance behavior, rather than rejecting a design solely because it is not orthogonal.

```{r, eval=FALSE}
a <- rsm_design_audit(fit)
a$fds
a$information_indices
rsm_plot(fit, type="prediction_variance")
```

When additional experimental runs are still possible, `rsm_augment()` searches a candidate region greedily for points that improve a D- or A-oriented information criterion without modifying observations already collected.

```{r, eval=FALSE}
rsm_augment(fit, n_add=4, objective="D")
```

This design-aware perspective follows the broader principle that response-surface design should be judged by inferential and predictive consequences over the region of interest [@BoxDraper1959].
