---
title: "rsmFlow: Classical RSM, Geometry, and Diagnostics"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{rsmFlow: Classical RSM, Geometry, and Diagnostics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse=TRUE, comment="#>", fig.width=7, fig.height=5, dpi=120, eval=FALSE)
```

# Classical RSM, geometry, and diagnostics

**Role in the vignette curriculum.** This vignette owns hierarchical polynomial fitting, residual and influence diagnostics, lack of fit, canonical analysis, steepest paths, and ridge exploration. Design construction is covered in the design vignette; bounded decision optimization and uncertainty are covered separately.

## 1. Classical RSM is a local modeling language

A second-order response surface is

\[
Y = \beta_0 + \sum_i \beta_i x_i + \sum_i \beta_{ii}x_i^2 + \sum_{i<j}\beta_{ij}x_i x_j + \varepsilon.
\]

The polynomial is especially useful for local exploration and optimization because its geometry is explicit. It should not automatically be interpreted as a mechanistic biological law.

The package follows the classical RSM tradition exemplified by `rsm` [@Lenth2009], while adding a design-aware and decision-aware workflow.

## 2. Load the agronomic teaching example

```{r}
library(rsmFlow)
p <- system.file("extdata", "agronomy_rsm_nonorthogonal.csv", package="rsmFlow")
dat <- read.csv(p)
head(dat)
```

## 3. First-order versus second-order models

### First order

```{r}
fo <- rsm_fit(
  dat,
  response="Yield",
  factors=c("N","K","Irrigation"),
  order=1
)
summary(fo)
```

A first-order model is appropriate for directional search when curvature is weak over the studied region.

### Second order

```{r}
so <- rsm_fit(
  dat,
  response="Yield",
  factors=c("N","K","Irrigation"),
  order=2
)
summary(so)
```

A second-order model adds quadratic curvature and pairwise interactions.

## 4. Hierarchy is deliberate

If a quadratic term \(x_i^2\) is retained, the corresponding linear term \(x_i\) should generally remain. If an interaction \(x_i x_j\) is retained, the main effects should generally remain.

This hierarchy makes the model interpretable under changes of origin and supports coherent surface geometry.

`rsm_fit(order=2)` constructs the hierarchical second-order structure automatically.

## 5. OLS, WLS, and robust coefficient estimation

Ordinary least squares:

```{r}
fit_ols <- rsm_fit(dat, "Yield", c("N","K","Irrigation"), order=2, estimator="ols")
```

Power-weighted least squares:

```{r}
fit_wls <- rsm_fit(dat, "Yield", c("N","K","Irrigation"), order=2, estimator="wls_power")
```

Robust coefficient estimation:

```{r}
fit_rob <- rsm_fit(dat, "Yield", c("N","K","Irrigation"), order=2, estimator="robust")
```

### Interpretation

Changing the estimator should solve a diagnosed problem, not serve as an automatic model-selection contest. Compare residual behavior, coefficient stability, and resulting decisions.

## 6. Blocking

When an experiment was blocked, encode that structure rather than treating the block effect as unexplained residual variation.

```{r}
# Example if a column named Block is present:
# fit_block <- rsm_fit(
#   dat_blocked,
#   response="Yield",
#   factors=c("N","K"),
#   order=2,
#   block="Block"
# )
```

The exact interpretation depends on the design. `rsmFlow` currently treats the block as part of the fixed model structure in this classical workflow; it is not a general mixed-model engine.

## 7. Diagnose before interpreting geometry

```{r}
d <- rsm_diagnostics(so)
d
```

Typical quantities include:

| Diagnostic | Interpretation |
|---|---|
| Residuals | unexplained response pattern |
| Standardized residuals | scale-adjusted departures |
| Leverage | unusual model-matrix positions |
| Cook's distance | influence on the fitted model |
| RMSE/MAE | descriptive fit error |
| R² | fraction of variation represented in Gaussian polynomial fit |
| Lack-of-fit result | model structure versus pure-error replication, when available |

## 8. Residual-versus-fitted plot

```{r}
rsm_plot(so, type="residuals")
```

Look for:

- curvature remaining in residuals;
- changing spread;
- groups or bands;
- isolated influential observations.

Residuals should not be treated as a decorative final figure. They can invalidate the interpretation of the fitted surface.

## 9. Normal Q-Q plot

```{r}
rsm_plot(so, type="qq")
```

Moderate tail deviations can occur by chance in small experiments. Combine the plot with the scientific scale, replication, and sensitivity analyses rather than treating a normality test as a binary gate.

## 10. Leverage and Cook's distance

```{r}
rsm_plot(so, type="leverage")
rsm_plot(so, type="cooks")
```

High leverage is common at axial or extreme design points because those runs define curvature. A high-leverage point is not automatically an error.

Investigate provenance before excluding any run.

## 11. Lack of fit and pure error

If identical factor combinations were replicated, pure error can be separated from model lack of fit.

```{r}
lof <- rsm_lack_of_fit(so)
lof
```

### Interpretation

A meaningful lack-of-fit signal means the current polynomial structure does not explain systematic variation beyond replicated experimental error.

Possible responses include:

- transform or change response family;
- expand the model only with scientific justification;
- fit a biologically motivated nonlinear model;
- inspect a missing block/batch structure;
- reconsider the experimental region.

Do not respond by optimizing the inadequate surface more aggressively.

## 12. Visualize the mean surface

For a three-factor model, fix one factor and view two at a time.

```{r}
rsm_plot(
  so,
  type="contour",
  x="N",
  y="K",
  at=list(Irrigation=median(dat$Irrigation))
)
```

```{r}
rsm_plot(
  so,
  type="surface",
  x="N",
  y="K",
  at=list(Irrigation=median(dat$Irrigation)),
  interactive=TRUE
)
```

Every conditional plot should state the fixed values of omitted factors.

## 13. Canonical analysis

A second-order surface can be written as

\[
\hat y = \beta_0 + \mathbf{b}^\top\mathbf{x} + \mathbf{x}^\top\mathbf{B}\mathbf{x}.
\]

If \(\mathbf B\) is nonsingular, the stationary point is

\[
\mathbf{x}_s = -\frac{1}{2}\mathbf{B}^{-1}\mathbf b.
\]

Run:

```{r}
can <- rsm_canonical(so)
can
```

Inspect:

```{r}
can$stationary_point
can$eigenvalues
can$nature
```

## 14. Classify the stationary point

| Eigenvalue pattern | Geometry |
|---|---|
| all negative | local maximum |
| all positive | local minimum |
| mixed signs | saddle |
| one or more near zero | ridge / nearly flat direction |

This classification describes the mathematical surface around the stationary point. It does not guarantee that the point lies inside the experimental bounds.

## 15. Plot canonical directions

```{r}
rsm_plot(so, type="canonical")
```

Canonical axes can reveal directions of strong and weak curvature that are hidden by the original factor axes.

## 16. A synthetic maximum for teaching

```{r}
set.seed(410)
N <- rep(seq(40,160,length.out=7), each=7)
P <- rep(seq(20,100,length.out=7), times=7)
Y <- 12 - 0.00025*(N-110)^2 - 0.00055*(P-62)^2 + rnorm(length(N),0,0.08)
max_dat <- data.frame(N,P,Y)
max_fit <- rsm_fit(max_dat,"Y",c("N","P"),order=2)
rsm_canonical(max_fit)
```

Expected interpretation: both principal curvatures should be negative and the stationary point should be near the designed peak.

## 17. A synthetic saddle for teaching

```{r}
set.seed(411)
x1 <- runif(60,-1,1)
x2 <- runif(60,-1,1)
y <- 5 + 2*x1^2 - 1.5*x2^2 + rnorm(60,0,0.08)
saddle_dat <- data.frame(x1,x2,y)
saddle_fit <- rsm_fit(saddle_dat,"y",c("x1","x2"),order=2)
rsm_canonical(saddle_fit)
```

A saddle is neither a maximum nor a minimum. Searching for “the optimum” without first understanding this geometry can be misleading.

## 18. Steepest ascent

A first-order model provides a gradient direction.

```{r}
fo2 <- rsm_fit(max_dat,"Y",c("N","P"),order=1)
path_up <- rsm_steepest(
  fo2,
  direction="ascent",
  step=0.10,
  n_steps=12,
  standardized=TRUE
)
path_up
```

### Interpretation

Steepest ascent is a sequential-experiment tool. The usual logic is:

1. fit a local first-order surface;
2. move along the estimated ascent direction;
3. conduct real runs along the path;
4. stop when response no longer improves;
5. center a new curvature experiment near the promising region.

Predictions along the path do not replace real sequential runs.

## 19. Steepest descent

```{r}
path_down <- rsm_steepest(
  fo2,
  direction="descent",
  step=0.10,
  n_steps=12
)
```

Useful when the scientific objective is minimization, such as disease, contamination, energy use, or residual concentration.

## 20. Ridge analysis

When the stationary point is outside the experimental region or curvature is weak, explore best response on expanding radii around a reference center.

```{r}
ridge <- rsm_ridge(
  so,
  radii=seq(0.1,1.0,by=0.1),
  goal="max",
  n_starts=30,
  seed=412
)
ridge
```

### Interpretation

Ridge analysis separates two questions:

- What is the local canonical geometry?
- What is the best fitted response at increasing distance from the center?

This is particularly useful for elongated or weakly curved surfaces.

## 21. Profiles and conditional interpretation

```{r}
prof <- rsm_profile(
  so,
  factor="N",
  at=list(
    K=median(dat$K),
    Irrigation=median(dat$Irrigation)
  ),
  n=100
)
head(prof)
```

```{r}
rsm_plot(
  so,
  type="profile",
  x="N",
  at=list(K=median(dat$K), Irrigation=median(dat$Irrigation))
)
```

A profile is conditional on the fixed values of all other factors. It should not be described as a marginal causal effect without additional design assumptions.

## 22. Perturbation plot

```{r}
rsm_perturbation(so)
rsm_plot(so, type="perturbation")
```

Perturbation curves compare local response sensitivity around a common reference point. They are not global variable-importance scores.

## 23. Model comparison: FO versus SO

Cross-validation can complement geometry and diagnostics.

```{r}
cv_fo <- rsm_cv(fo, folds=5, repeats=10, seed=413)
cv_so <- rsm_cv(so, folds=5, repeats=10, seed=413)
cv_fo
cv_so
```

If the second-order model does not improve predictive performance and curvature is weak, a simpler first-order representation can remain scientifically preferable.

## 24. Three interpretation scenarios

### Scenario A: interior maximum

Report stationary coordinates, negative eigenvalues, bounded confirmation, prediction precision, and uncertainty.

### Scenario B: stationary point outside the region

Do not extrapolate the mathematical maximum as an experimental recommendation. Continue with bounded optimization and possibly a new sequential design.

### Scenario C: saddle or ridge

Do not force a “maximum” narrative. Use ridge/search-path tools and reconsider the experimental region.

## 25. Common mistakes

1. Interpreting coefficients before checking hierarchy and design correlation.
2. Treating a statistically significant quadratic term as proof of an interior optimum.
3. Calling a saddle point an optimum.
4. Optimizing a model with clear lack of fit.
5. Removing axial runs simply because they have high leverage.
6. Reporting a conditional 2D slice as if all other factors varied freely.
7. Confusing canonical geometry with a constrained operational decision.

## 26. Minimum reporting checklist

- [ ] Model order and hierarchy reported.
- [ ] Factor coding/natural scale stated.
- [ ] Estimator identified.
- [ ] Block term reported where used.
- [ ] Residual diagnostics examined.
- [ ] Lack of fit reported when pure-error replication exists.
- [ ] Influential runs investigated rather than automatically deleted.
- [ ] Stationary point coordinates reported.
- [ ] Canonical eigenvalues/classification reported.
- [ ] Whether stationary point lies inside the experimental region stated.
- [ ] Conditional values reported for 2D slices of higher-dimensional models.
- [ ] Extrapolation clearly separated from interpolation.

## 27. Where to go next

If Gaussian polynomial assumptions are inappropriate or biological saturation is central, continue to **GLM and Nonlinear Response Surfaces**. If model adequacy is acceptable and the next question is a decision, continue to **Optimization, Uncertainty, and Economics**.
