---
title: "Optimization, uncertainty and near-optimal regions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Optimization, uncertainty and near-optimal regions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
---

`rsmFlow` distinguishes four objects that are often conflated in applied analyses: a stationary point, a bounded optimum, a global-search optimum, and an economic optimum.

```{r, eval=FALSE}
rsm_canonical(fit)
rsm_optimize(fit, method="L-BFGS-B")
rsm_optimize(fit, method="GA")
rsm_optimize(fit, method="hybrid")
```

Point estimates can be misleading when curvature is weak or the optimum is close to a boundary. The package therefore refits and reoptimizes bootstrap samples to estimate uncertainty in factor coordinates and the predicted response.

```{r, eval=FALSE}
opt <- rsm_optimize(fit, method="hybrid")
unc <- rsm_optimum_ci(fit, opt, B=1999, method="residual")
unc$intervals
```

Formal confidence-set methodology for optimal factor levels is a recognized inferential problem in response-surface analysis [@Wan2016]. The bootstrap tools here are intended as a transparent computational layer; exact or specialized confidence-region methods can be added through optional backends.

A practical decision often does not require the exact mathematical maximizer. Define a near-optimal region by

\[
\mathcal R_\delta=\{\mathbf x:\widehat y(\mathbf x)\ge \widehat y_{\max}-\delta|\widehat y_{\max}|\}.
\]

```{r, eval=FALSE}
rsm_near_optimal(fit, opt, tolerance=.05)
```

`rsm_robust_optimize()` can instead maximize a lower confidence bound, trading a small amount of fitted response for lower prediction uncertainty.
