---
title: "Multiple responses and economic optimum"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Multiple responses and economic optimum}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

For multiple responses, `rsmFlow` supports desirability and an approximate Pareto front. If \(d_j\) denotes an individual desirability and \(w_j\) an importance weight, the combined desirability is

\[
D=\exp\left\{\frac{\sum_j w_j\log(d_j)}{\sum_j w_j}\right\}.

This geometric-combination approach follows the classical desirability framework of Derringer and Suich [@DerringerSuich1980].
\]

```{r, eval=FALSE}
res <- rsm_multiopt(
  list(fit_yield, fit_quality),
  goals = c("max", "target"),
  limits = list(
    list(low=6, high=10),
    list(low=12, target=14, high=16)
  ),
  method="desirability"
)
```

The package also includes an economic-decision layer without becoming a general economics package. For constant response price \(p_y\), factor costs \(c_i\), and fixed cost \(C_0\), the objective is

\[
\Pi(\mathbf x)=p_y\widehat y(\mathbf x)-\sum_i c_i(x_i-x_{i,0})-C_0.
\]

```{r, eval=FALSE}
eco <- rsm_economic_optimum(
  fit,
  response_price = 1200,
  factor_cost = c(N=5.5, K=4.0, Irrigation=900),
  fixed_cost = 350
)
```

The returned object reports both the economic solution and the unconstrained biological maximum, making the agronomic trade-off explicit. Nonlinear revenue and factor-cost functions can be supplied as R functions.
