---
title: "rsmFlow: Optimization, Optimum Uncertainty, and Economics"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{rsmFlow: Optimization, Optimum Uncertainty, and Economics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse=TRUE, comment="#>", fig.width=7, fig.height=5, dpi=120, eval=FALSE)
```

# Optimization, optimum uncertainty, and economics

**Role in the vignette curriculum.** This deep dive owns constrained optimization, target search, optimizer comparison, uncertainty in optimum coordinates, joint regions, near-optimal operating zones, uncertainty-aware optimization, optimum stability, and biological-versus-economic decisions. Multiresponse trade-offs are developed in the next vignette.

## 1. An optimum is an estimated decision, not a fact

A response-surface optimum depends on:

- model specification;
- estimated parameters;
- factor bounds;
- optimization objective;
- optimizer behavior;
- uncertainty in the fitted surface.

Therefore, a defensible analysis reports more than one coordinate vector.

For optimum inference, uncertainty in the factor settings themselves is central [@Wan2016].

## 2. Load and fit the agronomic example

```{r}
library(rsmFlow)
p <- system.file("extdata","agronomy_rsm_nonorthogonal.csv",package="rsmFlow")
dat <- read.csv(p)
fit <- rsm_fit(dat,"Yield",c("N","K","Irrigation"),order=2)
rsm_diagnostics(fit)
```

Only continue if model adequacy is scientifically acceptable.

## 3. Stationary geometry versus bounded decision

```{r}
can <- rsm_canonical(fit)
can
```

The stationary point solves the unconstrained quadratic geometry. The operational decision is restricted to feasible bounds.

```{r}
opt <- rsm_optimize(
  fit,
  goal="max",
  method="hybrid",
  seed=601
)
opt
```

### Interpretation

If the bounded optimum lies on a factor limit, report that explicitly. A boundary optimum means the experiment supports a best point **within the tested region**, not necessarily a global biological maximum.

## 4. Explicit bounds

```{r}
b <- list(
  N=range(dat$N),
  K=range(dat$K),
  Irrigation=range(dat$Irrigation)
)

opt_b <- rsm_optimize(
  fit,
  goal="max",
  method="L-BFGS-B",
  bounds=b
)
```

Use tighter bounds when engineering, agronomic, regulatory, or safety constraints restrict the feasible operating region.

## 5. Grid search

```{r}
opt_grid <- rsm_optimize(
  fit,
  goal="max",
  method="grid",
  grid_n=41
)
```

Grid search is transparent and useful for teaching and verification, but resolution grows rapidly with dimensionality.

## 6. L-BFGS-B

```{r}
opt_lbfgs <- rsm_optimize(
  fit,
  goal="max",
  method="L-BFGS-B"
)
```

Efficient for smooth bounded surfaces.

## 7. Genetic algorithm

```{r}
opt_ga <- rsm_optimize(
  fit,
  goal="max",
  method="GA",
  seed=602,
  ga_control=list(popSize=100,maxiter=400,run=80)
)
```

Use a stochastic global search when multimodality or difficult constraints make local search less reliable. Freeze the seed for reproducibility.

## 8. Hybrid optimization

```{r}
opt_hybrid <- rsm_optimize(
  fit,
  goal="max",
  method="hybrid",
  seed=602
)
```

Hybrid optimization combines global exploration with local refinement.

## 9. Compare all methods

```{r}
all_opt <- rsm_optimize(
  fit,
  goal="max",
  method="all",
  seed=602
)
all_opt
```

### Interpretation

Agreement among methods on a smooth surface increases confidence in numerical optimization. Disagreement should trigger inspection of flat ridges, multiple local extrema, scaling, and bounds.

## 10. Minimize a response

```{r}
# Example for disease or contaminant concentration:
# rsm_optimize(disease_fit, goal="min", method="hybrid", seed=603)
```

The same framework applies to maximization and minimization.

## 11. Hit a target instead of max/min

Suppose a response should be near a quality target rather than as large as possible.

```{r}
opt_target <- rsm_optimize(
  fit,
  goal="target",
  target=85,
  method="hybrid",
  seed=604
)
```

Target optimization is useful for pH, firmness, concentration, moisture, and quality specifications.

## 12. Why optimum coordinates are uncertain

If fitted coefficients are random, the optimum is a derived random quantity:

\[
\hat{\mathbf{x}}_{opt}=h(\hat{\boldsymbol\beta}).
\]

A coefficient table does not directly quantify uncertainty in \(h(\cdot)\).

## 13. Bootstrap / coefficient-simulation optimum uncertainty

```{r}
unc <- rsm_optimum_ci(
  fit,
  optimum=opt_hybrid,
  B=999,
  conf=.95,
  optimizer="L-BFGS-B",
  seed=605
)
unc
```

Inspect successful and failed re-optimizations. Failed replicates are information about numerical or model instability and should not be silently discarded.

## 14. Plot the optimum cloud

```{r}
rsm_plot_optimum_uncertainty(
  unc,
  x="N",
  y="K",
  conf=.95
)
```

For three-factor models, plot multiple 2D projections or conditional summaries.

### Interpretation

An elongated cloud indicates correlated uncertainty in factor settings. A cloud concentrated on a boundary can indicate truncation by the feasible region rather than precise knowledge of an unconstrained biological optimum.

## 15. Joint optimum region

```{r}
reg <- rsm_optimum_region(
  unc,
  factors=c("N","K"),
  conf=.95,
  n=300
)
head(reg)
```

The current implementation is an approximate region based on the bootstrap/simulation distribution; it should not be described as an exact Fieller-type confidence set.

## 16. Near-optimal region

A practical operating zone can be more useful than a single point.

```{r}
near <- rsm_near_optimal(
  fit,
  optimum=opt_hybrid,
  tolerance=.05,
  goal="max",
  n=41,
  max_points=150000,
  seed=606
)
head(near)
```

For maximization, a 5% tolerance represents points with fitted response close to the best fitted response under the function's definition.

### Interpretation

Near-optimal regions can reveal opportunities to:

- reduce input without substantial yield loss;
- choose settings less sensitive to operational error;
- avoid poorly predicted boundaries;
- trade small response losses for economic or environmental benefits.

## 17. Optimum stability

```{r}
stab <- rsm_optimum_stability(
  unc,
  fit,
  tolerance=.10
)
stab
```

This asks whether uncertainty replicates stay within a practically acceptable neighborhood of the reference decision.

## 18. Uncertainty-aware optimization

```{r}
rob <- rsm_robust_optimize(
  fit,
  goal="max",
  lambda=1.96,
  method="L-BFGS-B",
  seed=607
)
rob
```

This penalizes poorly supported high predictions according to the package's uncertainty criterion.

### Interpretation

A conservative optimum can be preferable when the apparent peak is located in a region of high prediction uncertainty.

## 19. Visual comparison of fitted and conservative decisions

```{r}
rsm_plot(
  fit,
  type="contour",
  x="N",
  y="K",
  at=list(Irrigation=median(dat$Irrigation)),
  optimum=opt_hybrid
)
```

Add the uncertainty-aware point in a custom `ggplot2` layer when preparing publication figures.

## 20. Economic optimization

Biological maximum:

\[
\max_{\mathbf x}\hat y(\mathbf x).
\]

Economic objective:

\[
\max_{\mathbf x}\left[
P_y\hat y(\mathbf x)
-\sum_i c_i(x_i-x_{i,0})
-C_0
\right].
\]

Here \(x_{i,0}\) is a baseline at which incremental factor cost begins.

```{r}
eco <- rsm_economic_optimum(
  fit,
  response_price=12,
  factor_cost=c(
    N=0.80,
    K=0.55,
    Irrigation=35
  ),
  fixed_cost=120,
  baseline=c(
    N=min(dat$N),
    K=min(dat$K),
    Irrigation=min(dat$Irrigation)
  ),
  method="L-BFGS-B",
  seed=608
)
eco
```

## 21. Why baseline matters

Using incremental input above a baseline prevents nonsensical negative costs when factors are coded around zero. Economic assumptions should be stated in natural units.

## 22. Biological versus economic optimum

Report both:

```{r}
eco$biological_optimum
eco$solution
```

### Interpretation

The economic optimum can use less input than the biological maximum because the marginal response no longer pays for the marginal input.

This is a decision consequence, not evidence that the biological model was wrong.

## 23. Sensitivity to price assumptions

```{r}
prices <- c(8,10,12,15,20)

sens <- lapply(prices, function(py) {
  ans <- rsm_economic_optimum(
    fit,
    response_price=py,
    factor_cost=c(N=.80,K=.55,Irrigation=35),
    fixed_cost=120,
    baseline=c(N=min(dat$N),K=min(dat$K),Irrigation=min(dat$Irrigation)),
    method="L-BFGS-B"
  )
  cbind(data.frame(price=py), as.data.frame(as.list(ans$solution)), data.frame(profit=ans$profit))
})
do.call(rbind,sens)
```

### Interpretation

An economic optimum without sensitivity to uncertain prices can be less informative than a near-optimal economic region across plausible prices.

## 24. Boundary optima need explicit language

If any optimal coordinate equals its lower or upper bound, write:

> “The best fitted setting within the experimental region occurred on the upper N boundary.”

Do not write:

> “The true optimum N rate is the upper tested dose.”

The first is supported by the experiment. The second extrapolates beyond the evidence.

## 25. Uncertainty method depends on model family

For Gaussian polynomial models, coefficient simulation and bootstrap-style workflows are natural.

For likelihood GLMs, parametric simulate/refit/re-optimize is preferred where implemented.

For quasi families, asymptotic coefficient simulation is more appropriate than pretending a full likelihood exists.

For nonlinear models, refit bootstrap is preferable when the error model supports it; generalized nonlinear variance/correlation structures need additional care.

Always inspect the `method` stored in the uncertainty object.

## 26. Three decision scenarios

### Scenario A: sharp interior maximum

Expect a compact optimum cloud and relatively small near-optimal zone.

### Scenario B: broad plateau

Expect large coordinate uncertainty but potentially stable predicted response. Report an operating zone rather than one coordinate.

### Scenario C: boundary optimum

Expect truncation. Consider a new experiment beyond the current boundary only if scientifically feasible and safe.

## 27. Three complete code recipes

### Classical yield maximum

```{r}
fit <- rsm_fit(dat,"Yield",c("N","K","Irrigation"),order=2)
opt <- rsm_optimize(fit,goal="max",method="hybrid",seed=700)
unc <- rsm_optimum_ci(fit,optimum=opt,B=999,seed=700)
near <- rsm_near_optimal(fit,optimum=opt,tolerance=.05,seed=700)
```

### GLM pest minimum

```{r}
# pest_fit <- rsm_glm_fit(pest,"Count",c("Dose","Time"),family=poisson())
# pest_opt <- rsm_optimize(pest_fit,goal="min",scale="response")
# pest_unc <- rsm_optimum_ci(pest_fit,pest_opt,B=999,seed=701)
```

### Nonlinear fertilizer maximum

```{r}
# nl_fit <- rsm_nonlinear_fit(nd,"Yield",c("N","P"),
#   model="mitscherlich2_product",start="ga",seed=702)
# nl_opt <- rsm_optimize(nl_fit,goal="max",method="hybrid",seed=702)
# nl_unc <- rsm_optimum_ci(nl_fit,nl_opt,B=999,seed=702)
```

## 28. Common mistakes

1. Reporting the unconstrained stationary point as a feasible recommendation.
2. Ignoring boundary status.
3. Giving optimum coordinates without uncertainty.
4. Hiding failed bootstrap/refit replicates.
5. Treating a covariance ellipse approximation as an exact confidence region.
6. Reporting a single point when a broad near-optimal zone exists.
7. Optimizing a model that failed diagnostics.
8. Using economic costs on coded units without defining a natural-unit baseline.
9. Treating one price scenario as permanent truth.

## 29. Minimum reporting checklist

- [ ] Objective (`max`, `min`, `target`) stated.
- [ ] Factor bounds stated in natural units.
- [ ] Optimization method reported.
- [ ] Random seed reported for stochastic optimization.
- [ ] Boundary coordinates identified.
- [ ] Uncertainty method, B, and confidence level reported.
- [ ] Failed uncertainty replicates reported.
- [ ] Joint region described as approximate when appropriate.
- [ ] Near-optimal tolerance defined.
- [ ] Stability criterion defined.
- [ ] Economic price/cost units stated.
- [ ] Cost baseline stated.
- [ ] Biological and economic optima distinguished.
- [ ] Price/cost sensitivity considered when decisions depend strongly on them.

## 30. Where to go next

Continue to **Multiresponse Optimization and Tier 3** when several responses conflict or when simple response-surface engines do not adequately represent the predictive structure.
