---
title: "rsmFlow: Multiresponse Optimization and Tier-3 Surrogates"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{rsmFlow: Multiresponse Optimization and Tier-3 Surrogates}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse=TRUE, comment="#>", fig.width=7, fig.height=5, dpi=120, eval=TRUE)
```

# Multiresponse optimization and Tier-3 surrogates

**Role in the vignette curriculum.** This deep dive owns multiresponse desirability, weighted/distance/epsilon compromises, Pareto fronts and optional NSGA-II, advanced GP/GAM/TPS/RF/NN surrogate models, cross-validation, engine/optimum comparison, and one-step Bayesian optimization. Single-response uncertainty and economic optimization are developed elsewhere.

## 1. Why multiple responses change the decision problem

Many experiments have more than one scientifically relevant outcome. A fertilizer combination can increase yield while reducing quality or increasing disease. A processing condition can maximize extraction while also increasing energy use. A formulation can improve efficacy while moving a quality attribute away from its target.

In such problems there is usually no single response surface whose maximum solves the entire decision.

`rsmFlow` supports several transparent compromise strategies:

- individual and overall desirability [@DerringerSuich1980];
- weighted compromise;
- distance-to-target;
- epsilon constraints;
- Pareto fronts;
- optional NSGA-II search.

## 2. Create a realistic three-response teaching experiment

```{r}
library(rsmFlow)
set.seed(801)
md <- expand.grid(
  N=seq(40,160,length.out=7),
  Irrigation=seq(0.55,1.25,length.out=7)
)

md$Yield <- 9.5 + 0.035*md$N + 4.0*md$Irrigation -
  0.00014*md$N^2 - 1.5*md$Irrigation^2 +
  0.008*md$N*md$Irrigation + rnorm(nrow(md),0,.10)

md$Quality <- 80 + 0.035*md$N + 2.2*md$Irrigation -
  0.00018*(md$N-100)^2 - 2.8*(md$Irrigation-.85)^2 +
  rnorm(nrow(md),0,.15)

md$Disease <- 18 - 0.030*md$N + 8.5*(md$Irrigation-.65) +
  0.00010*(md$N-90)^2 + rnorm(nrow(md),0,.18)
```

Fit one response model per outcome:

```{r}
fit_y <- rsm_fit(md,"Yield",c("N","Irrigation"),order=2)
fit_q <- rsm_fit(md,"Quality",c("N","Irrigation"),order=2)
fit_d <- rsm_fit(md,"Disease",c("N","Irrigation"),order=2)
```

Before combining them, diagnose each model separately.

```{r}
rsm_diagnostics(fit_y)
rsm_diagnostics(fit_q)
rsm_diagnostics(fit_d)
```

A multiresponse compromise cannot rescue an inadequate individual response model.

## 3. Individual desirability

Desirability maps each response to \([0,1]\), where 0 is unacceptable and 1 is most desirable under the declared preference rule.

### Maximize yield

```{r}
yhat <- predict(fit_y, md)
d_y <- rsm_desirability(
  yhat,
  goal="max",
  low=quantile(yhat,.10),
  high=quantile(yhat,.90),
  weight=1
)
```

### Minimize disease

```{r}
dhat <- predict(fit_d, md)
d_d <- rsm_desirability(
  dhat,
  goal="min",
  low=quantile(dhat,.10),
  high=quantile(dhat,.90),
  weight=1
)
```

### Hit a quality target

```{r}
qhat <- predict(fit_q, md)
d_q <- rsm_desirability(
  qhat,
  goal="target",
  low=80,
  target=83,
  high=85,
  weight=1
)
```

### Interpretation

Desirability is a decision transformation, not a probability and not an inferential statistic. The limits and targets must be scientifically justified.

## 4. Overall desirability

```{r, eval=FALSE}
limits <- list(
  list(low=10, high=14),
  list(low=80, target=83, high=85),
  list(low=8, high=18)
)

des_opt <- rsm_multiopt(
  fits=list(fit_y,fit_q,fit_d),
  goals=c("max","target","min"),
  limits=limits,
  method="desirability",
  importance=c(1,1,1),
  optimizer="GA",
  seed=802
)
des_opt
```

### Interpretation

The solution is conditional on the desirability specification. Report the limits, targets, weights, and final individual desirabilities together with the overall score.

## 5. Visualize the desirability surface

```{r, eval=FALSE}
dsurf <- rsm_desirability_surface(
  fits=list(fit_y,fit_q,fit_d),
  goals=c("max","target","min"),
  limits=limits,
  importance=c(1,1,1),
  x="N",
  y="Irrigation",
  n=100
)
dsurf
```

The surface can reveal whether the compromise is sharp or whether a broad desirable operating zone exists.

## 6. Weighted compromise

```{r, eval=FALSE}
weighted <- rsm_multiopt(
  fits=list(fit_y,fit_q,fit_d),
  goals=c("max","max","min"),
  limits=limits,
  method="weighted",
  importance=c(2,1,1.5),
  optimizer="GA",
  seed=803
)
weighted
```

Weights encode analyst or stakeholder priorities. Sensitivity analysis is essential.

## 7. Distance-to-target

```{r, eval=FALSE}
distance <- rsm_multiopt(
  fits=list(fit_y,fit_q),
  goals=c("target","target"),
  limits=list(
    list(low=10,target=13,high=14),
    list(low=80,target=83,high=85)
  ),
  method="distance",
  importance=c(1,1),
  seed=804
)
distance
```

Useful when several responses have explicit engineering or quality targets.

## 8. Epsilon constraints

Suppose yield is the primary objective and disease must stay below a hard limit.

```{r, eval=FALSE}
eps <- rsm_multiopt(
  fits=list(fit_y,fit_d),
  goals=c("max","min"),
  limits=list(
    list(low=10,high=14),
    list(low=8,high=18)
  ),
  method="epsilon",
  primary=1,
  epsilon=list(NULL, 12),
  seed=805
)
eps
```

Epsilon constraints are often preferable when one response represents a mandatory safety, quality, or environmental threshold rather than a symmetric preference.

## 9. Pareto front by sampling

```{r, eval=FALSE}
pareto_sample <- rsm_pareto(
  fits=list(fit_y,fit_d),
  goals=c("max","min"),
  n=10000,
  engine="sample",
  seed=806
)
head(pareto_sample)
```

Plot:

```{r, eval=FALSE}
rsm_plot_pareto(
  pareto_sample,
  response_x="response1",
  response_y="response2"
)
```

### Interpretation

A Pareto point is non-dominated. Improving one response from that point requires worsening at least one other response.

A Pareto front presents trade-offs; it does not choose a final policy for the analyst.

## 10. Optional NSGA-II

```{r, eval=FALSE}
if (requireNamespace("mco", quietly=TRUE)) {
  pareto_nsga <- rsm_pareto(
    fits=list(fit_y,fit_d),
    goals=c("max","min"),
    engine="NSGA2",
    seed=807
  )
  pareto_nsga
}
```

NSGA-II is appropriate when continuous multiobjective search is preferable to a finite random candidate cloud. Backend behavior must be checked under the installed `mco` version.

## 11. Overlaid constraints

```{r}
over <- rsm_overlaid_contour(
  fits=list(fit_y,fit_q,fit_d),
  constraints=list(
    list(goal="max", low=12.5),
    list(goal="max", low=82),
    list(goal="min", high=12)
  ),
  x="N",
  y="Irrigation",
  n=120
)
over
```

This asks a different question from desirability: **where are all hard requirements satisfied simultaneously?**

# Part II. Tier-3 surrogate modeling

## 12. Why a Tier-3 layer exists

Classical RSM is interpretable and efficient. GLM and nonlinear models extend it when distribution or biological shape requires it. Tier-3 surrogates should be used only when these simpler models do not adequately represent the predictive problem.

Available methods:

```{r}
# methods accepted by rsm_surrogate():
# "quadratic", "gam", "gp", "tps", "rf", "nn"
```

The package does not automatically declare a Tier-3 model superior because training error is lower.

## 13. Generate a nonlinear teaching surface

```{r}
set.seed(810)
sd <- expand.grid(
  x1=seq(-1,1,length.out=11),
  x2=seq(-1,1,length.out=11)
)
sd$y <- 4 + sin(pi*sd$x1)*cos(pi*sd$x2) +
  0.6*sd$x1*sd$x2 + rnorm(nrow(sd),0,.08)
```

## 14. Quadratic benchmark

```{r}
quad <- rsm_surrogate(
  sd,"y",c("x1","x2"),
  method="quadratic",
  seed=811
)
```

A simple benchmark is essential. Advanced methods should demonstrate value relative to an interpretable baseline.

## 15. GAM

```{r}
if (requireNamespace("mgcv", quietly=TRUE)) {
  gam_fit <- rsm_surrogate(
    sd,"y",c("x1","x2"),
    method="gam",
    seed=811
  )
}
```

The GAM implementation uses smooth factor effects and interaction structure suitable for flexible response surfaces.

## 16. Thin-plate spline

```{r}
if (requireNamespace("mgcv", quietly=TRUE)) {
  tps_fit <- rsm_surrogate(
    sd,"y",c("x1","x2"),
    method="tps",
    seed=811
  )
}
```

TPS provides a joint smooth surface rather than a purely additive interpretation.

## 17. Gaussian Process

```{r}
if (requireNamespace("DiceKriging", quietly=TRUE)) {
  gp_fit <- rsm_surrogate(
    sd,"y",c("x1","x2"),
    method="gp",
    seed=811
  )
}
```

Gaussian Processes are especially useful when predictive uncertainty is needed for sequential experimentation [@JonesSchonlauWelch1998].

## 18. Random Forest

```{r}
if (requireNamespace("ranger", quietly=TRUE)) {
  rf_fit <- rsm_surrogate(
    sd,"y",c("x1","x2"),
    method="rf",
    seed=811
  )
}
```

RF is flexible but generally less transparent as a smooth response-surface equation.

## 19. Neural network

```{r}
if (requireNamespace("nnet", quietly=TRUE)) {
  nn_fit <- rsm_surrogate(
    sd,"y",c("x1","x2"),
    method="nn",
    seed=811
  )
}
```

Neural networks can approximate complex surfaces but require careful scaling, tuning, and validation.

## 20. Surrogate optimization

```{r}
if (exists("gp_fit")) {
  gp_opt <- rsm_surrogate_optimize(
    gp_fit,
    goal="max",
    method="L-BFGS-B",
    seed=812
  )
  gp_opt
}
```

A flexible predictor can generate a numerically sharp optimum even when predictive uncertainty is high. Do not separate optimization from validation.

## 21. Cross-validation

For `rsmFlow_fit` objects:

```{r}
classical <- rsm_fit(sd,"y",c("x1","x2"),order=2)
cv_classical <- rsm_cv(classical,folds=5,repeats=10,seed=813)
cv_classical
```

For Tier-3 engines, use backend-aware validation and `rsm_compare_engines()` summaries where available.

## 22. Compare engines without automatic ranking

```{r, eval=FALSE}
models <- list(classical=classical)
if (exists("gam_fit")) models$gam <- gam_fit
if (exists("gp_fit")) models$gp <- gp_fit
if (exists("rf_fit")) models$rf <- rf_fit
if (exists("nn_fit")) models$nn <- nn_fit

do.call(rsm_compare_engines, models)
```

Interpret separate dimensions:

| Dimension | Question |
|---|---|
| Fit error | Does the model reproduce the training data? |
| CV error | Does it generalize to held-out data? |
| Diagnostics | Are residual/model assumptions defensible? |
| Interpretability | Can the factor-response structure be explained? |
| Optimum stability | Is the decision sensitive to model choice? |
| Uncertainty | Is predictive uncertainty available/calibrated? |
| Complexity | Is the added flexibility scientifically justified? |

## 23. Compare optima directly

```{r}
opt_classical <- rsm_optimize(classical,goal="max",method="hybrid",seed=814)

if (exists("gp_fit")) {
  opt_gp <- rsm_surrogate_optimize(gp_fit,goal="max",method="L-BFGS-B",seed=814)
  rsm_compare_optima(opt_classical,opt_gp)
}
```

Similar RMSE can coexist with materially different optima. Decision comparison is therefore its own validation layer.

## 24. One-step Bayesian optimization

Expected Improvement combines predicted mean and uncertainty to propose a new candidate [@JonesSchonlauWelch1998].

```{r}
if (requireNamespace("DiceKriging", quietly=TRUE)) {
  next_run <- rsm_bayes_opt(
    sd,
    response="y",
    factors=c("x1","x2"),
    goal="max",
    n_candidates=10000,
    xi=.01,
    seed=815
  )
  next_run
}
```

### Critical interpretation

`rsm_bayes_opt()` proposes the next experimental setting. It does **not** generate a new observed response and does not automatically run a closed experimental loop.

Correct workflow:

1. fit the surrogate;
2. calculate acquisition;
3. propose the next setting;
4. conduct the real experiment;
5. measure the response;
6. append the observation;
7. refit and repeat if justified.

## 25. Exploration versus exploitation

The Expected Improvement parameter `xi` changes the balance between exploiting high predicted regions and exploring uncertainty.

```{r}
# Compare proposals:
# rsm_bayes_opt(sd,"y",c("x1","x2"),xi=.001,seed=816)
# rsm_bayes_opt(sd,"y",c("x1","x2"),xi=.05,seed=816)
```

Document acquisition settings in a sequential-design report.

## 26. Three escalation scenarios

### Scenario A: quadratic model works well

Stay with classical RSM. Interpretability and established geometry are advantages.

### Scenario B: smooth but strongly nonlinear surface

Consider GAM/TPS or a biologically motivated nonlinear model before a black-box learner.

### Scenario C: expensive experiment with sequential runs

A GP with calibrated predictive uncertainty and Bayesian optimization can be valuable.

## 27. Common mistakes

1. Combining inadequate response models into a multiresponse score.
2. Treating desirability thresholds as data-derived facts when they are policy choices.
3. Reporting one Pareto solution without the trade-off front.
4. Using NSGA-II when a simple grid already solves the low-dimensional problem transparently.
5. Ranking Tier-3 models by in-sample RMSE.
6. Ignoring optimum disagreement among models.
7. Treating GP uncertainty as automatically calibrated.
8. Treating a Bayesian-optimization proposal as an observed result.
9. Adding an advanced model without a simpler benchmark.

## 28. Minimum reporting checklist

- [ ] Every response model diagnosed separately.
- [ ] Goal for each response stated.
- [ ] Desirability low/target/high values reported.
- [ ] Importance weights reported and sensitivity explored.
- [ ] Hard constraints distinguished from preferences.
- [ ] Pareto definition and search method reported.
- [ ] NSGA-II backend/version recorded when used.
- [ ] Tier-3 method and tuning settings reported.
- [ ] Validation method reported.
- [ ] Simpler benchmark included.
- [ ] Optima compared across plausible engines.
- [ ] Bayesian acquisition function/parameters/seed reported.
- [ ] Proposed next run clearly distinguished from a measured response.

## 29. Where to go next

Continue to **Visualization, Shiny, and Teaching** for a coherent communication layer, then to **Validation, Reproducibility, and Method Selection** for release-quality scientific verification.
