---
title: "Generalized response surfaces: GLM-RSM for two quantitative factors"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Generalized response surfaces: GLM-RSM}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse=TRUE,comment="#>")
library(rsmFlow)
```

## Why a GLM response surface?

For two quantitative factors, `rsm_glm_fit()` fits a first- or second-order predictor

\[
g(\mu)=\beta_0+\beta_1x_1+\beta_2x_2+\beta_{11}x_1^2+\beta_{22}x_2^2+\beta_{12}x_1x_2,
\]

while predictions and operational optimization are reported on the response scale by default. This is useful when the response support or mean-variance relationship is incompatible with ordinary Gaussian least squares.

The GLM module never changes family automatically. Dispersion, residuals, lack of fit, scientific support, and sensitivity alternatives should be considered together.

## Example 1: Poisson surface for insect counts

```{r}
set.seed(101)
d <- expand.grid(Dose=seq(-1,1,length.out=7), Days=seq(-1,1,length.out=7))
mu <- exp(2.2 + .30*d$Dose - .18*d$Days - .35*d$Dose^2 - .20*d$Days^2 + .10*d$Dose*d$Days)
d$Insects <- rpois(nrow(d),mu)
fit_p <- rsm_glm_fit(d,"Insects",c("Dose","Days"),family=poisson(link="log"))
rsm_glm_dispersion(fit_p)
opt_p <- rsm_optimize(fit_p,"min","L-BFGS-B")
opt_p
```

The dispersion diagnostic is a screening tool rather than an automatic model selector. If substantial extra-Poisson variability is present, a negative-binomial or quasi-Poisson sensitivity analysis may be scientifically justified.

## Example 2: binomial response using trial totals

```{r}
set.seed(102)
d$Trials <- 40
eta <- -0.4 + .9*d$Dose - .35*d$Days - .55*d$Dose^2 + .12*d$Dose*d$Days
prob <- plogis(eta)
d$Success <- rbinom(nrow(d),d$Trials,prob)
d$Proportion <- d$Success/d$Trials
fit_b <- rsm_glm_fit(d,"Proportion",c("Dose","Days"),family=binomial(),weights=d$Trials)
rsm_glm_diagnostics(fit_b)$dispersion
rsm_plot(fit_b,"contour")
```

## Example 3: Gamma surface for a positive asymmetric response

```{r}
set.seed(103)
mean_y <- exp(1.6 + .15*d$Dose + .18*d$Days - .22*d$Dose^2 - .15*d$Days^2)
d$Positive <- rgamma(nrow(d),shape=10,scale=mean_y/10)
fit_g <- rsm_glm_fit(d,"Positive",c("Dose","Days"),family=Gamma(link="log"))
ca_g <- rsm_glm_canonical(fit_g)
ca_g
rsm_plot(fit_g,"heatmap")
```

## Exposure offsets

When counts represent unequal exposures, `offset=` is encoded as an explicit offset term so the same semantics are valid for `glm` and `MASS::glm.nb`. `offset_reference=` defines the link-scale exposure used for new response-surface predictions and optimization. For a log exposure, `offset_reference=0` represents unit exposure.

```{r}
set.seed(104)
o <- expand.grid(Dose=seq(-1,1,length.out=6),Days=seq(-1,1,length.out=6))
o$exposure <- rep(c(.5,1,2),length.out=nrow(o))
rate <- exp(1.7-.25*o$Dose^2-.15*o$Days^2)
o$Count <- rpois(nrow(o),o$exposure*rate)
fit_off <- rsm_glm_fit(o,"Count",c("Dose","Days"),family=poisson(),
  offset=log(o$exposure),offset_reference=0)
rsm_optimize(fit_off,"max","L-BFGS-B")
```

## Negative binomial

When `MASS` is available, `family="negative_binomial"` routes to `MASS::glm.nb()`. The estimated dispersion parameter is retained in the backend model. GLM-RSM optimization still occurs on the response scale.

```{r eval=requireNamespace("MASS",quietly=TRUE)}
set.seed(104)
d$Overdispersed <- MASS::rnegbin(nrow(d),mu=exp(2-.3*d$Dose^2-.2*d$Days^2),theta=2)
fit_nb <- rsm_glm_fit(d,"Overdispersed",c("Dose","Days"),family="negative_binomial")
summary(fit_nb)
```

## Optimum uncertainty

For likelihood-based GLM surfaces the default uncertainty workflow is parametric: simulate a response under the fitted family, refit the whole response surface, reoptimize, and retain the optimum coordinates. This respects the model-specific mean-variance relation better than treating GLM residuals as homoscedastic Gaussian errors. Quasi-Poisson and quasi-binomial models do not define a unique full sampling likelihood, so `rsmFlow` uses asymptotic coefficient simulation for optimum uncertainty rather than pretending that a unique parametric bootstrap distribution exists.

```{r eval=FALSE}
unc <- rsm_optimum_ci(fit_p,optimum=opt_p,B=999,method="parametric")
rsm_optimum_region(unc)
rsm_optimum_stability(unc,fit_p)
```
