---
title: "rsmFlow: Visualization, Shiny, and Teaching Workflows"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{rsmFlow: Visualization, Shiny, and Teaching Workflows}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse=TRUE, comment="#>", fig.width=7, fig.height=5, dpi=120, eval=TRUE)
```

# Visualization, Shiny, and teaching workflows

**Role in the vignette curriculum.** This vignette owns communication: static and interactive graphics, higher-dimensional views, the optional Shiny interface, GUI-to-code reproducibility, and teaching sequences. Statistical model details are intentionally referenced rather than repeated.

## 1. Every plot should answer a question

A response surface can be visualized in many ways, but each view emphasizes different evidence.

| Plot | Best question |
|---|---|
| Design plot | Where were runs placed? |
| Term-correlation plot | Which model terms are related? |
| Prediction-variance map | Where are predictions weak? |
| Contour | Where do fitted response levels occur? |
| Heat map | How does response change over a 2D region? |
| 3D surface | What is the overall fitted geometry? |
| Profile | How does one factor change response conditionally? |
| Perturbation | Which factor directions are locally influential? |
| Slice | How does a 2D surface change as a third factor is fixed? |
| FDS / VDG | How is precision distributed over the region? |
| Optimum cloud | How uncertain are factor settings? |
| Pareto plot | What trade-offs exist among responses? |

## 2. Fit the standard agronomic example

```{r}
library(rsmFlow)
p <- system.file("extdata","agronomy_rsm_nonorthogonal.csv",package="rsmFlow")
dat <- read.csv(p)
fit <- rsm_fit(dat,"Yield",c("N","K","Irrigation"),order=2)
```

## 3. Show the design first

```{r}
rsm_plot(fit,type="design")
```

Observed design points provide essential context for every fitted surface. A smooth prediction region can otherwise look more informative than the actual data support.

## 4. Term correlation

```{r}
rsm_plot(fit,type="term_correlation")
```

Use this with the numerical audit rather than as a decorative heat map.

## 5. Contour plot

```{r}
rsm_plot(
  fit,
  type="contour",
  x="N",
  y="K",
  at=list(Irrigation=median(dat$Irrigation)),
  show_points=TRUE
)
```

Always report the fixed third-factor value in the caption.

## 6. Heat map

```{r}
rsm_plot(
  fit,
  type="heatmap",
  x="N",
  y="K",
  at=list(Irrigation=median(dat$Irrigation))
)
```

Useful for communication to readers who find contour geometry unfamiliar.

## 7. 3D response surface

```{r}
rsm_plot(
  fit,
  type="surface",
  x="N",
  y="K",
  at=list(Irrigation=median(dat$Irrigation)),
  interactive=FALSE
)
```

Interactive Plotly version:

```{r}
rsm_plot(
  fit,
  type="surface",
  x="N",
  y="K",
  at=list(Irrigation=median(dat$Irrigation)),
  interactive=TRUE
)
```

3D surfaces are excellent for exploration but can distort slope perception. Pair them with contours for precise interpretation.

## 8. Prediction variance as a figure

```{r}
rsm_plot(fit,type="prediction_variance",x="N",y="K")
```

A publication showing an optimum can benefit from a companion precision map, especially for non-orthogonal or augmented designs.

## 9. Residual diagnostic figures

```{r}
rsm_plot(fit,type="residuals")
rsm_plot(fit,type="qq")
rsm_plot(fit,type="leverage")
rsm_plot(fit,type="cooks")
```

These belong before the optimization figure in the analytical logic, even if they appear in supplementary material in a manuscript.

## 10. Profiles

```{r}
prof_N <- rsm_profile(
  fit,
  factor="N",
  at=list(
    K=median(dat$K),
    Irrigation=median(dat$Irrigation)
  )
)
head(prof_N)
```

```{r}
rsm_plot(
  fit,
  type="profile",
  x="N",
  at=list(K=median(dat$K),Irrigation=median(dat$Irrigation))
)
```

A profile is conditional. State the held-constant settings.

## 11. Perturbation

```{r}
rsm_perturbation(fit)
rsm_plot(fit,type="perturbation")
```

Use for local sensitivity around a common reference point. Do not call it a global importance ranking.

## 12. Sliced surfaces for higher-dimensional RSM

```{r}
slices <- rsm_slices(
  fit,
  x="N",
  y="K",
  slice_factor="Irrigation",
  values=quantile(dat$Irrigation,c(.25,.5,.75)),
  n=50
)
slices
```

This is the correct visual grammar for a three-factor model: a sequence of conditional two-factor surfaces.

## 13. Overlaid constraints

For multiple responses:

```{r}
# rsm_overlaid_contour(
#   fits=list(fit_yield,fit_quality,fit_disease),
#   constraints=list(
#     list(goal="max",low=12),
#     list(goal="max",low=82),
#     list(goal="min",high=12)
#   ),
#   x="N",y="Irrigation"
# )
```

Show the feasible operating region rather than only the selected compromise point.

## 14. Desirability surface

```{r}
# rsm_desirability_surface(
#   fits=list(fit1,fit2),
#   goals=c("max","target"),
#   limits=list(
#     list(low=10,high=14),
#     list(low=80,target=83,high=85)
#   )
# )
```

## 15. Pareto figure

```{r}
# p <- rsm_pareto(list(fit1,fit2),c("max","min"),n=10000,seed=901)
# rsm_plot_pareto(p,response_x="response1",response_y="response2")
```

A Pareto plot should highlight non-dominated solutions and label response directions clearly.

## 16. Optimum uncertainty figure

```{r, eval=FALSE}
opt <- rsm_optimize(fit,goal="max",method="hybrid",seed=902)
unc <- rsm_optimum_ci(fit,optimum=opt,B=999,seed=902)
rsm_plot_optimum_uncertainty(unc,x="N",y="K")
```

The cloud communicates coordinate uncertainty more directly than three separate interval lines.

## 17. FDS and VDG for design teaching

```{r}
rsm_plot(fit,type="fds")
rsm_plot(fit,type="vdg")
```

These figures belong to design-quality teaching rather than to generic residual diagnostics.

## 18. A publication figure sequence

For a two-factor optimization paper, a coherent figure set can be:

1. design geometry with observed runs;
2. residual/diagnostic panel;
3. fitted contour with bounded optimum;
4. prediction-variance map;
5. optimum uncertainty cloud or near-optimal region;
6. biological-versus-economic comparison if costs matter.

The sequence mirrors the evidence chain.

# Part II. Shiny as an interactive teaching layer

## 19. Launch the app

```{r, eval=FALSE}
run_rsm_app()
```

The Shiny interface is optional. The scientific library remains usable without it.

## 20. The GUI is not a second statistical engine

The design principle is:

```text
Shiny controls
    ↓
public rsmFlow functions
    ↓
standard result objects
    ↓
plots / tables / reports
```

This keeps script and GUI analyses aligned.

## 21. GUI-to-code bridge: classical RSM

GUI path:

```text
Import data
→ choose response
→ choose quantitative factors
→ second-order model
→ fit
```

Equivalent R:

```{r}
fit <- rsm_fit(
  dat,
  response="Yield",
  factors=c("N","K","Irrigation"),
  order=2
)
```

## 22. GUI-to-code bridge: design audit

Equivalent R:

```{r}
rsm_design_audit(fit)
rsm_fds(fit)
rsm_vdg(fit)
```

Students should learn to copy the conceptual sequence from the GUI into a reproducible script.

## 23. GUI-to-code bridge: optimization

Equivalent R:

```{r, eval=FALSE}
opt <- rsm_optimize(fit,goal="max",method="hybrid",seed=903)
unc <- rsm_optimum_ci(fit,optimum=opt,B=999,seed=903)
near <- rsm_near_optimal(fit,optimum=opt,tolerance=.05,seed=903)
```

## 24. GUI-to-code bridge: GLM

```{r}
# fit_glm <- rsm_glm_fit(
#   count_data,"Count",c("Dose","Time"),family=poisson()
# )
# rsm_glm_diagnostics(fit_glm)
# rsm_optimize(fit_glm,goal="min",scale="response")
```

## 25. GUI-to-code bridge: nonlinear model

```{r}
# fit_nl <- rsm_nonlinear_fit(
#   data,"Yield",c("N","P"),
#   model="mitscherlich2_product",
#   start="ga",seed=904
# )
```

## 26. Teaching sequence for beginners

A useful laboratory class can follow:

### Stage 1: inspect the runs

Students identify response, factors, units, bounds, center points, and replication.

### Stage 2: fit but do not optimize

Students fit FO/SO models and inspect residuals.

### Stage 3: understand geometry

Students classify maximum/minimum/saddle/ridge examples.

### Stage 4: bounded decision

Students compare stationary and bounded optima.

### Stage 5: uncertainty

Students inspect an optimum cloud and near-optimal zone.

### Stage 6: write the R script equivalent

The GUI becomes a conceptual scaffold rather than a black box.

## 27. Teaching sequence for advanced users

1. Audit a deliberately non-orthogonal design.
2. Compare prediction variance, FDS, and VDG.
3. Compare Gaussian RSM with GLM/nonlinear alternatives.
4. Freeze seeds and compare optimizers.
5. Build a desirability compromise and Pareto front.
6. Compare classical RSM with a GP/GAM surrogate.
7. Propose a Bayesian-optimization run without fabricating its response.
8. Produce a scientific report with explicit limitations.

## 28. Use `rsm_explain()` as a teaching aid

```{r}
rsm_explain(fit)
```

Use explanatory output to support interpretation, not to replace statistical reasoning.

## 29. Figure interpretation prompts for students

For every contour plot ask:

- Where are the observed points?
- Which region is interpolated?
- Is the apparent optimum inside the design?
- Is the response flat in any direction?
- How would prediction variance change the recommendation?

For every uncertainty cloud ask:

- Is uncertainty symmetric?
- Is it truncated by bounds?
- Would multiple settings be practically equivalent?

For every Pareto plot ask:

- Which responses conflict?
- Which solutions are dominated?
- What additional value judgment is required to choose one point?

## 30. Communication principles

1. Keep natural units on final axes whenever possible.
2. Show observed points with fitted surfaces.
3. State fixed factor values on slices.
4. Distinguish response and link scales.
5. Label boundary optima explicitly.
6. Report uncertainty around decisions.
7. Avoid decorative 3D figures when a contour communicates the result more precisely.
8. Use color plus shape/line type when accessibility matters.

## 31. Common mistakes

1. Showing a surface without observed points.
2. Omitting the fixed third-factor value.
3. Using interactive graphics as the only reproducible record.
4. Teaching the GUI without teaching the equivalent R function.
5. Calling an optimum cloud a confidence ellipse without checking the underlying method.
6. Using a Pareto plot without explaining dominance.
7. Hiding diagnostic warnings because the final contour looks smooth.

## 32. Where to go next

Continue to **Validation, Reproducibility, and Method Selection** before treating a model or interface workflow as release-ready scientific software.
