---
title: "rsmFlow: Validation, Reproducibility, and Method Selection"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{rsmFlow: Validation, Reproducibility, and Method Selection}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse=TRUE, comment="#>", fig.width=7, fig.height=5, dpi=120, eval=FALSE)
```

# Validation, reproducibility, and method selection

**Role in the vignette curriculum.** This final deep dive owns validation logic, model comparison, simulation design, frozen seeds, backend checks, reproducibility, state-of-the-art positioning, and release-quality evidence. It does not repeat detailed tutorials for individual models.

## 1. Validation has several meanings

A package can parse successfully and still be scientifically wrong. A model can fit successfully and still be inappropriate. A numerical method can return a solution and still be unstable.

Separate at least five validation layers:

1. **Software validity:** functions run and return correctly structured objects.
2. **Numerical validity:** results agree with trusted calculations or known truths.
3. **Statistical validity:** estimators/inference behave as intended under controlled scenarios.
4. **Scientific validity:** model assumptions and decisions make sense for the experimental problem.
5. **Documentation validity:** examples describe the actual implemented API and behavior.

## 2. Current development-snapshot policy

The vignette sources use `eval=FALSE` because the generation environment for this documentation reorganization does not certify a complete R runtime and all optional specialist backends.

Therefore:

- syntax and API references can be audited statically here;
- runtime execution belongs to the local validation workflow;
- no vignette result should be treated as runtime-certified merely because the Rmd file exists.

This distinction should remain visible until `R CMD check --as-cran` and the scientific test battery pass in a complete environment.

## 3. Start with the official smoke test

From the package root:

```{r}
source("scripts/SMOKE_TEST.R")
```

The smoke test covers representative classical, design-audit, canonical, optimization, economic, GLM, nonlinear, and Tier-3 paths when optional dependencies are available.

A smoke test is a breadth gate, not proof of numerical correctness.

## 4. Run automated tests

```{r}
if (requireNamespace("testthat", quietly=TRUE)) {
  testthat::test_dir("tests/testthat")
}
```

The test suite should include both success paths and failure/guardrail paths.

Examples of scientifically consequential guardrails:

- mixture designs rejected because they are outside scope;
- target optimization requires a target;
- insufficient rank reported;
- GLM family/scale rules enforced;
- invalid nonlinear starts rejected;
- failed bootstrap refits reported;
- epsilon constraints checked;
- optional backends fail informatively when absent.

## 5. R CMD build and check

Canonical release gates:

```text
R CMD build rsmFlow
R CMD check --as-cran rsmFlow_<version>.tar.gz
```

A static source ZIP is not a substitute for the tarball produced by `R CMD build`.

## 6. Freeze stochastic experiments

Every simulation scenario should have:

- a scenario identifier;
- a fixed data-generating mechanism;
- fixed factor region;
- fixed model specification;
- fixed seed sequence;
- fixed number of replicates;
- predefined metrics;
- predefined failure rules.

Do not change seeds or tolerances after observing inconvenient results without versioning the simulation protocol.

## 7. Simulation battery structure

The package includes `scripts/SIMULATION_BATTERY.R`. The current intended battery covers:

| Scenario | Question |
|---|---|
| A | design geometry × non-orthogonality × noise |
| B | interior, boundary, and external stationary points |
| C | optimum interval coverage |
| D | heteroscedasticity: OLS vs power-WLS |
| E | block and lack-of-fit behavior |
| F | scaling from 2 to multiple factors |
| G | design augmentation under correlation |
| H | conflicting multiresponse decisions |
| I | biological vs economic optimum |
| J | GLM Poisson/binomial/Gamma behavior |
| K | nonlinear parameter recovery |
| L | GA vs multistart initialization |
| M | Tier-3 surrogate comparison |
| N | generalized nonlinear heteroscedasticity |

The scenario list should remain frozen for comparative reruns within a release cycle.

## 8. Validate design audit numerically

Use known model matrices where rank and condition are analytically or independently computable.

```{r}
set.seed(1001)
d <- data.frame(
  x1=runif(60,-1,1),
  x2=runif(60,-1,1)
)
d$y <- 3 + d$x1 - .5*d$x2 + .8*d$x1^2 + rnorm(60,0,.05)
fit <- rsm_fit(d,"y",c("x1","x2"),order=2)
a <- rsm_design_audit(fit)
a
```

Compare rank and condition diagnostics with independent linear-algebra calculations where appropriate.

## 9. Validate canonical analysis with known surfaces

### Known maximum

```{r}
set.seed(1002)
dmax <- expand.grid(x1=seq(-1,1,length.out=9),x2=seq(-1,1,length.out=9))
dmax$y <- 10 - 2*(dmax$x1-.2)^2 - 1.5*(dmax$x2+.1)^2 + rnorm(nrow(dmax),0,.02)
fmax <- rsm_fit(dmax,"y",c("x1","x2"),order=2)
rsm_canonical(fmax)
```

Expected: maximum near `(0.2, -0.1)` with negative eigenvalues.

### Known saddle

```{r}
set.seed(1003)
ds <- expand.grid(x1=seq(-1,1,length.out=9),x2=seq(-1,1,length.out=9))
ds$y <- 5 + 2*ds$x1^2 - 1.5*ds$x2^2 + rnorm(nrow(ds),0,.02)
fs <- rsm_fit(ds,"y",c("x1","x2"),order=2)
rsm_canonical(fs)
```

Expected: mixed eigenvalue signs.

## 10. Validate bounded optimization against known truth

```{r}
truth <- c(x1=.2,x2=-.1)
ans <- rsm_optimize(fmax,goal="max",method="all",seed=1004)
ans
```

Compare analytic/grid/local/global/hybrid solutions within a predefined tolerance.

## 11. Validate boundary behavior

Construct a surface whose unconstrained maximum is outside `[-1,1]^2`.

```{r}
set.seed(1005)
db <- expand.grid(x1=seq(-1,1,length.out=9),x2=seq(-1,1,length.out=9))
db$y <- 10 - (db$x1-1.5)^2 - (db$x2-.2)^2 + rnorm(nrow(db),0,.02)
fb <- rsm_fit(db,"y",c("x1","x2"),order=2)
rsm_canonical(fb)
rsm_optimize(fb,goal="max",method="hybrid",seed=1005)
```

Expected: stationary point outside the region and bounded optimum on the `x1=1` edge.

## 12. Optimum uncertainty coverage

For repeated simulated datasets with known interior optimum:

1. generate data;
2. fit model;
3. obtain optimum interval/region;
4. record whether the true optimum is covered;
5. summarize coverage across replicates.

A nominal 95% procedure should be evaluated by empirical coverage, not one attractive interval.

## 13. Heteroscedasticity

Generate variance increasing with fitted mean and compare:

```{r}
# fit_ols <- rsm_fit(dat,"y",c("x1","x2"),estimator="ols")
# fit_wls <- rsm_fit(dat,"y",c("x1","x2"),estimator="wls_power")
```

Compare coefficient bias, prediction error, interval behavior, and optimum stability.

Do not evaluate WLS only by lower residual SSE.

## 14. GLM simulation validation

For Poisson, binomial, and Gamma scenarios, freeze true parameters and assess:

- coefficient recovery;
- response-scale prediction error;
- dispersion behavior;
- bounded optimum recovery;
- uncertainty coverage;
- failure rate.

For quasi families, exclude likelihood AIC from methods that require a full likelihood.

## 15. Nonlinear parameter recovery

Use known Mitscherlich and Michaelis-Menten surfaces.

Metrics:

- convergence rate;
- parameter bias;
- RMSE;
- Jacobian condition;
- optimum bias;
- failed refit frequency.

A nonlinear model can predict well but estimate individual parameters poorly. Both dimensions matter.

## 16. GA versus multistart

Use identical parameter bounds and frozen datasets.

Compare:

| Metric | Why it matters |
|---|---|
| convergence frequency | practical reliability |
| final objective | quality of final estimator start |
| parameter recovery | scientific interpretation |
| runtime | computational cost |
| multiple minima | identifiability/search complexity |

Do not compare GA to multistart using different allowed parameter regions.

## 17. Tier-3 validation

The surrogate benchmark should use held-out or cross-validated performance. In-sample RMSE is insufficient for a claim of predictive superiority.

Recommended metrics:

- repeated CV RMSE/MAE;
- calibration of GP uncertainty where relevant;
- optimum bias relative to known truth;
- optimum variability across training samples;
- sensitivity to tuning;
- computational time.

## 18. One-step Bayesian optimization validation

Use synthetic functions with known global maxima. At each simulated iteration:

1. fit GP to observed runs;
2. propose candidate by Expected Improvement;
3. evaluate the **known synthetic truth** at that candidate;
4. append the true synthetic response;
5. repeat.

This is appropriate in simulation because the data-generating function is known. In a real experiment, step 3 must be a real measurement.

## 19. Model comparison is multi-criterion

```{r}
rsm_compare_engines(model1,model2,model3)
```

Do not collapse all evidence to one number.

For example:

| Criterion | Classical RSM | Nonlinear | GP |
|---|---|---|---|
| interpretation | strong | biological parameters | weaker/global |
| local geometry | explicit | model-specific | numerical |
| uncertainty | standard/bootstrap | refit/asymptotic | predictive GP |
| extrapolation | polynomial risk | biological form risk | usually weak |
| optimization | straightforward | straightforward | numerical |
| sequential design | limited | limited | strong with acquisition |

## 20. Compare decisions, not only models

```{r}
rsm_compare_optima(opt1,opt2,opt3)
```

A model comparison is scientifically incomplete when two plausible engines produce materially different recommended settings.

## 21. State-of-the-art positioning

`rsmFlow` is designed to complement rather than replace established RSM packages.

Relevant foundations include:

- classical RSM implementation and teaching in `rsm` [@Lenth2009];
- response-surface design principles [@BoxDraper1959];
- design selection under variance inflation considerations [@ODriscoll2015];
- confidence sets for optimal factor levels [@Wan2016];
- multiple-response desirability [@DerringerSuich1980];
- efficient global optimization and Expected Improvement [@JonesSchonlauWelch1998].

The package's intended contribution is integration: design-aware diagnostics, imperfect designs, classical/GLM/nonlinear surfaces, optimum uncertainty, economics, multiresponse decisions, Tier-3 surrogates, visualization, and teaching in one workflow.

Use bounded language. Do not claim that no other package can perform any individual component.

## 22. Reproducibility record

Every stochastic analysis should preserve:

```text
package version
R version
platform
optional backend versions
seed
factor bounds
optimizer settings
bootstrap B
candidate grids
model formula/family
```

Capture:

```{r}
sessionInfo()
```

For package development also preserve `DESCRIPTION`, `NAMESPACE`, test logs, and `00check.log`.

## 23. Dependency-aware validation

Optional packages currently used by specialist modules include examples such as:

```text
rsm
GA
MASS
minpack.lm
nlme
mgcv
DiceKriging
ranger
nnet
mco
plotly
ggplot2
OptimaRegion
vdg
AlgDesign
```

Not every suggested package is required for the core workflow. Test each optional backend explicitly if release documentation claims support for it.

## 24. Shiny validation

The optional Shiny interface should be tested against the same API:

1. import data;
2. fit classical model;
3. compare output with direct `rsm_fit()`;
4. run diagnostics;
5. optimize;
6. compare coordinates with direct `rsm_optimize()`;
7. export/report if enabled;
8. verify informative errors when optional packages are absent.

The GUI should never silently change the statistical method.

## 25. Documentation validation

Every exported function should have:

- an Rd page;
- valid argument documentation;
- realistic examples;
- consistent names across vignettes;
- no stale function calls.

The long vignettes should be statically audited for exported symbols before runtime rendering.

## 26. Local release sequence

Recommended order:

```text
parse R files
→ install dependencies
→ source smoke test
→ run testthat
→ run examples
→ render vignettes
→ run simulation battery
→ inspect numerical results
→ launch Shiny
→ R CMD build
→ R CMD check --as-cran
→ clean-library install
```

Do not report full validation before the relevant gates have actually run.

## 27. Failure classification

### Infrastructure failure

Examples: R not found, compiler unavailable, system library missing.

### Packaging failure

Examples: malformed DESCRIPTION, missing export, broken Rd, missing file.

### Runtime failure

Examples: function error, wrong class, backend API incompatibility.

### Numerical failure

Examples: optimizer converges to wrong known solution, unstable predictions, wrong Hessian sign.

### Methodological failure

Examples: interval undercoverage, design criterion not matching its documented definition, biased nonlinear parameter recovery.

The remedy depends on the class of failure.

## 28. Three model-selection case studies

### Case A: Gaussian yield data, adequate SO model

Prefer classical RSM unless an alternative meaningfully improves prediction or decision stability.

### Case B: overdispersed pest counts

Compare Poisson and Negative Binomial using dispersion, diagnostics, prediction, and response-scale optimum.

### Case C: saturating nutrient response

Compare quadratic and Mitscherlich models using diagnostics, biological parameter plausibility, cross-validation, and optimum location.

## 29. A minimum evidence table for a manuscript

| Domain | Evidence to report |
|---|---|
| Design | region, replication, audit/precision |
| Model | family/equation/order and hierarchy |
| Diagnostics | residual/dispersion/identifiability |
| Geometry | canonical class where relevant |
| Decision | bounded optimum and boundary status |
| Uncertainty | method, B, intervals/region |
| Alternatives | sensitivity/model comparison |
| Economics/multiresponse | assumptions/preferences |
| Reproducibility | version, seed, backend |

## 30. Common validation mistakes

1. Treating package installation as scientific validation.
2. Changing simulation seeds after seeing poor performance.
3. Reporting only successful bootstrap replicates without failures.
4. Ranking models by one information criterion.
5. Using in-sample Tier-3 error as predictive evidence.
6. Claiming exact I-optimal or confidence-region behavior without backend/numerical verification.
7. Ignoring version-specific optional backend APIs.
8. Calling static documentation checks runtime validation.
9. Assuming Shiny output is correct because the interface opens.

## 31. Final release checklist for the vignette ecosystem

- [ ] Eight final Rmd vignettes present.
- [ ] Old redundant short Rmd files removed from `vignettes/`.
- [ ] Shared `references.bib` retained.
- [ ] `VIGNETTE_ORGANIZATION.md` present.
- [ ] Every public function assigned to a primary vignette owner.
- [ ] No unresolved exported-symbol references.
- [ ] No missing citation keys.
- [ ] Code fences balanced.
- [ ] Runtime examples executed locally before release.
- [ ] `R CMD check --as-cran` passed before distribution claims.

## 32. Closing perspective

The purpose of validation is not to accumulate check marks. It is to establish that the sequence

**design -> model -> diagnosis -> optimization -> uncertainty -> decision**

behaves correctly under conditions where the truth or expected behavior is known, and remains transparent when the data are imperfect.

That evidence is what allows `rsmFlow` to function simultaneously as a teaching package and an advanced scientific workflow.
