---
title: "Tier 3: advanced surrogates, generalized nonlinear least squares, and Bayesian optimization"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tier 3 advanced response-surface models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse=TRUE,comment="#>")
library(rsmFlow)
set.seed(301)
d <- data.frame(x1=runif(70,-1,1),x2=runif(70,-1,1))
d$y <- sin(2.4*d$x1)+.7*cos(2*d$x2)+.25*d$x1*d$x2+rnorm(70,0,.05)
```

## Role of Tier 3

Tier 3 is explicit expert functionality. It is intended for scientifically justified model inadequacy, expensive experiments, complex smooth surfaces, or heterogeneous nonlinear errors. It is not an automatic replacement for the interpretable second-order response surface.

## Example 1: GAM/thin-plate surrogate

`method="gam"` uses factor-specific thin-plate smooths plus pairwise tensor interactions; `method="tps"` uses a joint multivariate thin-plate smooth. Both are explicit alternatives rather than automatic replacements for the quadratic RSM.


```{r eval=requireNamespace("mgcv",quietly=TRUE)}
gam_fit <- rsm_surrogate(d,"y",c("x1","x2"),method="gam")
rsm_surrogate_optimize(gam_fit,"max")
```

A thin-plate variant can be selected with `method="tps"`. Both are fitted with `mgcv` and remain outside the classical canonical-analysis workflow.

## Example 2: Gaussian process and expected-improvement recommendation

```{r eval=requireNamespace("DiceKriging",quietly=TRUE)}
gp <- rsm_surrogate(d,"y",c("x1","x2"),method="gp")
pr <- predict(gp,d[1:5,c("x1","x2")],se.fit=TRUE)
next_run <- rsm_bayes_opt(d,"y",c("x1","x2"),goal="max",n_candidates=1000)
next_run$next_point
```

`rsm_bayes_opt()` proposes one next experimental point from the current data. It never fabricates that future response and therefore cannot iterate without the user supplying the new observation.

## Example 3: random forest and neural network sensitivity models

```{r eval=requireNamespace("ranger",quietly=TRUE)}
rf <- rsm_surrogate(d,"y",c("x1","x2"),method="rf")
rsm_surrogate_optimize(rf,"max")
```

```{r eval=requireNamespace("nnet",quietly=TRUE)}
nn <- rsm_surrogate(d,"y",c("x1","x2"),method="nn",size=5,decay=.01)
rsm_surrogate_optimize(nn,"max")
```

These models are primarily predictive/sensitivity surrogates. Their use should be reported separately from polynomial canonical interpretation.

## Generalized nonlinear least squares as Tier 3

For mechanistic nonlinear surfaces with heterogeneous or correlated error, `engine="gnls"` provides a separate route through `nlme`. The mean structure remains explicit, while variance or correlation structures are declared by the analyst.
