---
title: "Comparing polynomial, GLM, nonlinear, and Tier-3 engines"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Comparing response-surface engines}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse=TRUE,comment="#>")
library(rsmFlow)
```

Model comparison in `rsmFlow` is deliberately multi-criterion. AIC, BIC, RMSE, convergence, diagnostics, scientific plausibility, prediction, and optimum stability answer different questions. Quasi-likelihood fits are not assigned likelihood AIC/BIC rankings by `rsm_compare_engines()`.

## Example 1: Gaussian LM versus Gaussian GLM

```{r}
set.seed(401)
d <- expand.grid(x1=seq(-1,1,length.out=8),x2=seq(-1,1,length.out=8))
d$y <- 8 + .4*d$x1-.3*d$x2-.7*d$x1^2-.4*d$x2^2+rnorm(nrow(d),0,.08)
lm_fit <- rsm_fit(d,"y",c("x1","x2"))
glm_fit <- rsm_glm_fit(d,"y",c("x1","x2"),family=gaussian())
rsm_compare_engines(LM=lm_fit,GLM=glm_fit)
```

## Example 2: polynomial approximation versus nonlinear saturation

```{r}
n <- expand.grid(N=seq(10,160,length.out=8),P=seq(5,90,length.out=7))
n$y <- 11*n$N/(30+n$N)*n$P/(20+n$P)+rnorm(nrow(n),0,.08)
poly <- rsm_fit(n,"y",c("N","P"))
nonlin <- rsm_nonlinear_fit(n,"y",c("N","P"),"michaelis_menten2",
  start=c(A=11,K1=30,K2=20),engine="nls")
rsm_compare_engines(Polynomial=poly,Nonlinear=nonlin)
```

## Example 3: compare operational optima

```{r}
op1 <- rsm_optimize(poly,"max","L-BFGS-B")
op2 <- rsm_optimize(nonlin,"max","L-BFGS-B")
rsm_compare_optima(Polynomial=op1,Nonlinear=op2)
```

A close optimum can coexist with different parameter interpretations; conversely, similar RMSE can hide materially different optima. For decision-oriented RSM, model adequacy should therefore be evaluated not only at the coefficient level but also in the location, uncertainty, and practical flatness of the optimum.
