---
title: "rsmFlow in Practice"
subtitle: "From Classical RSM to Advanced Surrogates"
author: "Walter Esfrain Pereira"
date: last-modified
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    number_sections: true
vignette: >
  %\VignetteIndexEntry{rsmFlow in Practice}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r library}
library(rsmFlow)
packageVersion("rsmFlow")
```

# Introduction {.unnumbered}

::: {.callout-note}
## What is rsmFlow?

The `rsmFlow` package provides a **design-aware response-surface workflow** for teaching and advanced scientific use.
:::

This tutorial walks through **five scientific scenarios**:

1. **Classical RSM** — polynomial surface, canonical analysis, optimization
2. **Optimization and uncertainty** — near-optimal regions, economic optimum
3. **GLM-RSM** — Poisson, Binomial, Gamma responses
4. **Nonlinear surfaces** — Mitscherlich, Michaelis-Menten with GA initialization
5. **Advanced surrogates** — GP, GAM, random forest, neural network, Bayesian optimization


# Part I: Classical RSM {#sec-classical}

## I.1 Fitting the model

```{r classic-fit}
set.seed(2026)
d_classic <- expand.grid(
  N = seq(80, 160, length.out = 6),
  K = seq(40, 120, length.out = 6)
)
d_classic$Yield <- with(d_classic,
  6 + 0.04 * (N - 120) - 0.0003 * (N - 120)^2 +
  0.03 * (K - 80) - 0.0002 * (K - 80)^2 +
  0.0001 * (N - 120) * (K - 80) + rnorm(nrow(d_classic), 0, 0.3)
)
fit <- rsm_fit(d_classic, "Yield", c("N", "K"), order = 2)
fit
```

## I.2 Design audit

```{r classic-audit}
rsm_design_audit(fit)
```

## I.3 Canonical analysis

```{r classic-canonical}
rsm_canonical(fit)
```

## I.4 Diagnostics

```{r classic-diagnostics, fig.cap="Diagnostic plots."}
rsm_diagnostics(fit)
```


# Part II: Optimization {#sec-optimization}

## II.1 Finding the optimum

```{r opt-optimize}
opt <- rsm_optimize(fit, goal = "max", method = "L-BFGS-B")
opt
```

## II.2 Near-optimal region

```{r opt-near}
near <- rsm_near_optimal(fit, opt, tolerance = 0.05, n = 20)
cat("Points in near-optimal region:", nrow(near), "\n")
```

## II.3 Economic optimum

```{r opt-economic}
eco <- rsm_economic_optimum(fit, response_price = 1200,
  factor_cost = c(N = 5.5, K = 4.0))
eco
```


# Part III: GLM-RSM {#sec-glm}

## III.1 Poisson

```{r glm-poisson}
set.seed(20260813)
g_pois <- expand.grid(
  Dose = seq(-1, 1, length.out = 7),
  Time = seq(-1, 1, length.out = 7)
)
eta_pois <- with(g_pois, 2 + 0.3 * Dose - 0.2 * Time -
  0.4 * Dose^2 - 0.25 * Time^2 + 0.15 * Dose * Time)
g_pois$count <- rpois(nrow(g_pois), lambda = exp(eta_pois))
fit_pois <- rsm_glm_fit(g_pois, "count", c("Dose", "Time"), family = poisson())
fit_pois
```

```{r glm-poisson-disp}
rsm_glm_dispersion(fit_pois)
```

## III.2 Binomial

```{r glm-binomial}
set.seed(20260814)
g_bin <- expand.grid(
  Dose = seq(-1, 1, length.out = 7),
  Time = seq(-1, 1, length.out = 7)
)
eta_bin <- with(g_bin, 0.5 + 0.8 * Dose - 0.4 * Time -
  0.6 * Dose^2 - 0.3 * Time^2 + 0.2 * Dose * Time)
g_bin$n <- 40
g_bin$success <- rbinom(nrow(g_bin), size = g_bin$n, prob = plogis(eta_bin))
g_bin$prop <- g_bin$success / g_bin$n
fit_bin <- rsm_glm_fit(g_bin, "prop", c("Dose", "Time"),
                       family = binomial(), weights = g_bin$n)
fit_bin
```

## III.3 Gamma

```{r glm-gamma}
set.seed(20260815)
g_gam <- expand.grid(
  x1 = seq(-1, 1, length.out = 7),
  x2 = seq(-1, 1, length.out = 7)
)
eta_gam <- with(g_gam, 1.8 + 0.25 * x1 - 0.20 * x2 -
  0.30 * x1^2 - 0.25 * x2^2)
g_gam$y <- rgamma(nrow(g_gam), shape = 10, scale = exp(eta_gam) / 10)
fit_gamma <- rsm_glm_fit(g_gam, "y", c("x1", "x2"),
                         family = Gamma(link = "log"))
fit_gamma
```


# Part IV: Nonlinear {#sec-nonlinear}

## IV.1 GA initialization

```{r nl-ga}
set.seed(20260816)
d_nl <- expand.grid(
  N = seq(10, 160, length.out = 7),
  P = seq(5, 90, length.out = 7)
)
d_nl$Yield <- with(d_nl,
  11 * (1 - exp(-0.025 * N)) * (1 - exp(-0.040 * P)) +
  rnorm(nrow(d_nl), 0, 0.08)
)
fit_ga <- rsm_nonlinear_fit(
  d_nl, "Yield", c("N", "P"),
  model = "mitscherlich2_product",
  start = "ga",
  start_control = list(popSize = 50, maxiter = 150, run = 50),
  engine = "nlsLM"
)
fit_ga
```

## IV.2 Diagnostics

```{r nl-diag}
rsm_nonlinear_diagnostics(fit_ga)
```

## IV.3 Michaelis-Menten

```{r nl-mm}
d_mm <- expand.grid(
  N = seq(10, 150, length.out = 6),
  P = seq(5, 80, length.out = 6)
)
d_mm$Yield <- with(d_mm, 10 * N / (30 + N) * P / (20 + P))
fit_mm <- rsm_nonlinear_fit(
  d_mm, "Yield", c("N", "P"),
  model = "michaelis_menten2",
  start = c(A = 10, K1 = 30, K2 = 20),
  engine = "nls"
)
rsm_nonlinear_diagnostics(fit_mm)
```


# Part V: Tier 3 Surrogates {#sec-tier3}

## V.1 Gaussian Process

```{r tier3-gp}
set.seed(1301)
d_gp <- data.frame(x1 = runif(50, -2, 2), x2 = runif(50, -2, 2))
d_gp$y <- with(d_gp, sin(x1) * cos(x2) + 0.3 * rnorm(nrow(d_gp)))
gp <- rsm_surrogate(d_gp, "y", c("x1", "x2"), method = "gp")
gp
```

## V.2 GAM

```{r tier3-gam}
gam_fit <- rsm_surrogate(d_gp, "y", c("x1", "x2"), method = "gam")
gam_fit
```

## V.3 Random Forest

```{r tier3-rf}
rf_fit <- rsm_surrogate(d_gp, "y", c("x1", "x2"), method = "rf")
pr_rf <- predict(rf_fit, d_gp[1:5, c("x1", "x2")])
cat("RF predictions:", round(pr_rf, 3), "\n")
```

## V.4 Neural Network

```{r tier3-nn}
nn_fit <- rsm_surrogate(d_gp, "y", c("x1", "x2"), method = "nn", size = 3)
pr_nn <- predict(nn_fit, d_gp[1:5, c("x1", "x2")])
cat("NN predictions:", round(pr_nn, 3), "\n")
```

## V.5 Bayesian Optimization

```{r tier3-bayes}
next_point <- rsm_bayes_opt(d_gp, "y", c("x1", "x2"), n_candidates = 1000)
cat("Recommended next point:\n")
print(next_point$next_point)
cat("Expected improvement:", round(next_point$expected_improvement, 4), "\n")
```


# Part VI: Graphics {#sec-graphics}

## VI.1 Perturbation plot

```{r fig-perturbation, fig.cap="Perturbation plot."}
rsm_perturbation(fit, center = opt$solution)
```

## VI.2 FDS plot

```{r fig-fds, fig.cap="Fraction of Design Space."}
rsm_fds(fit)
```

## VI.3 VDG

```{r fig-vdg, fig.cap="Variance Dispersion Graph."}
rsm_vdg(fit)
```


# References {.unnumbered}

- Box, G. E. P. & Draper, N. R. (2007). *Response Surfaces*. Wiley.
- Myers, R. H. et al. (2016). *Response Surface Methodology*. Wiley.
- Rasmussen, C. E. & Williams, C. K. I. (2006). *Gaussian Processes for Machine Learning*. MIT Press.

```{r session-info}
sessionInfo()
```
