---
title: "rsmFlow in Practice"
subtitle: "From Classical RSM to Advanced Surrogates"
author: "Walter Esfrain Pereira"
date: last-modified
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    number_sections: true
    fig_width: 7
    fig_height: 5
vignette: >
  %\VignetteIndexEntry{rsmFlow in Practice}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  out.width = "85%"
)
```

```{r library}
library(rsmFlow)
packageVersion("rsmFlow")
```

# Introduction {.unnumbered}

::: {.callout-note}
## What is rsmFlow?

The `rsmFlow` package provides a **design-aware response-surface workflow** for teaching and advanced scientific use.
:::

This tutorial walks through **five scientific scenarios**:

1. **Classical RSM** — polynomial surface, canonical analysis, optimization
2. **Optimization and uncertainty** — near-optimal regions, economic optimum
3. **GLM-RSM** — Poisson, Binomial, Gamma responses
4. **Nonlinear surfaces** — Mitscherlich, Michaelis-Menten with GA initialization
5. **Advanced surrogates** — GP, GAM, random forest, neural network, Bayesian optimization


# Part I: Classical RSM {#sec-classical}

## I.1 Data and fitting

```{r classic-data}
set.seed(2026)
d_classic <- expand.grid(
  N = seq(80, 160, length.out = 6),
  K = seq(40, 120, length.out = 6)
)
d_classic$Yield <- with(d_classic,
  6 + 0.04 * (N - 120) - 0.0003 * (N - 120)^2 +
  0.03 * (K - 80) - 0.0002 * (K - 80)^2 +
  0.0001 * (N - 120) * (K - 80) + rnorm(nrow(d_classic), 0, 0.3)
)
head(d_classic, 10)
```

```{r classic-scatter, fig.cap="Observed yield vs. nitrogen dose.", fig.alt="Scatterplot of yield vs N."}
plot(d_classic$N, d_classic$Yield, pch = 19, col = "#0072B2",
     xlab = "N (kg/ha)", ylab = "Yield (Mg/ha)",
     main = "Observed Data")
```

```{r classic-fit}
fit <- rsm_fit(d_classic, "Yield", c("N", "K"), order = 2)
fit
```

## I.2 Design audit

```{r classic-audit}
rsm_design_audit(fit)
```

## I.3 Canonical analysis

```{r classic-canonical}
rsm_canonical(fit)
```

## I.4 Diagnostics

```{r classic-diagnostics, fig.cap="Diagnostic plots for the classical RSM fit.", fig.alt="Four diagnostic panels."}
rsm_diagnostics(fit)
```

## I.5 Lack of fit

```{r classic-lof}
rsm_lack_of_fit(fit)
```


# Part II: Optimization {#sec-optimization}

## II.1 Finding the optimum

```{r opt-optimize}
opt <- rsm_optimize(fit, goal = "max", method = "L-BFGS-B")
opt
```

## II.2 Near-optimal region

```{r opt-near}
near <- rsm_near_optimal(fit, opt, tolerance = 0.05, n = 20)
cat("Points in near-optimal region:", nrow(near), "\n")
head(near, 10)
```

```{r opt-near-fig, fig.cap="Near-optimal region: points within 5% of the maximum.", fig.alt="Contour with near-optimal points."}
n1 <- seq(min(d_classic$N), max(d_classic$N), length.out = 50)
k1 <- seq(min(d_classic$K), max(d_classic$K), length.out = 50)
grd <- expand.grid(N = n1, K = k1)
grd$pred <- as.numeric(predict(fit, grd))
filled.contour(n1, k1, matrix(grd$pred, 50, 50),
               main = "Response Surface with Near-Optimal Region",
               xlab = "N (kg/ha)", ylab = "K (kg/ha)",
               color.palette = heat.colors)
points(near$N, near$K, pch = 19, col = rgb(0, 0, 0, 0.3), cex = 0.5)
points(opt$solution["N"], opt$solution["K"], pch = 18, col = "blue", cex = 1.5)
```

## II.3 Economic optimum

```{r opt-economic}
eco <- rsm_economic_optimum(fit, response_price = 1200,
  factor_cost = c(N = 5.5, K = 4.0))
eco
```

```{r opt-economic-fig, fig.cap="Biological vs economic optimum.", fig.alt="Contour with two optima."}
filled.contour(n1, k1, matrix(grd$pred, 50, 50),
               main = "Biological vs Economic Optimum",
               xlab = "N (kg/ha)", ylab = "K (kg/ha)",
               color.palette = heat.colors)
points(opt$solution["N"], opt$solution["K"], pch = 18, col = "red", cex = 1.5)
points(eco$solution["N"], eco$solution["K"], pch = 17, col = "blue", cex = 1.5)
legend("topright", legend = c("Biological", "Economic"),
       pch = c(18, 17), col = c("red", "blue"))
```


# Part III: GLM-RSM {#sec-glm}

## III.1 Poisson: count data

```{r glm-poisson}
set.seed(20260813)
g_pois <- expand.grid(
  Dose = seq(-1, 1, length.out = 7),
  Time = seq(-1, 1, length.out = 7)
)
eta_pois <- with(g_pois, 2 + 0.3 * Dose - 0.2 * Time -
  0.4 * Dose^2 - 0.25 * Time^2 + 0.15 * Dose * Time)
g_pois$count <- rpois(nrow(g_pois), lambda = exp(eta_pois))
fit_pois <- rsm_glm_fit(g_pois, "count", c("Dose", "Time"), family = poisson())
summary(fit_pois$model)$coefficients
```

```{r glm-poisson-disp}
rsm_glm_dispersion(fit_pois)
```

```{r glm-poisson-fig, fig.cap="Observed counts vs fitted values for Poisson GLM.", fig.alt="Scatterplot of observed vs fitted."}
plot(fitted(fit_pois$model), g_pois$count,
     pch = 19, col = "#D55E00",
     xlab = "Fitted values", ylab = "Observed counts",
     main = "Poisson GLM: Observed vs Fitted")
abline(0, 1, lty = 2, col = "grey50")
```

```{r glm-poisson-surface, fig.cap="Response surface on the link scale.", fig.alt="Contour plot of GLM predictions."}
opt_pois <- rsm_optimize(fit_pois, goal = "max", method = "L-BFGS-B")
nd_pois <- expand.grid(Dose = seq(-1, 1, length.out = 50), Time = seq(-1, 1, length.out = 50))
nd_pois$pred <- as.numeric(predict(fit_pois, nd_pois))
filled.contour(seq(-1, 1, length.out = 50), seq(-1, 1, length.out = 50),
               matrix(nd_pois$pred, 50, 50),
               main = "Poisson GLM Response Surface",
               xlab = "Dose", ylab = "Time",
               color.palette = heat.colors)
```

## III.2 Binomial: proportion data

```{r glm-binomial}
set.seed(20260814)
g_bin <- expand.grid(
  Dose = seq(-1, 1, length.out = 7),
  Time = seq(-1, 1, length.out = 7)
)
eta_bin <- with(g_bin, 0.5 + 0.8 * Dose - 0.4 * Time -
  0.6 * Dose^2 - 0.3 * Time^2 + 0.2 * Dose * Time)
g_bin$n <- 40
g_bin$success <- rbinom(nrow(g_bin), size = g_bin$n, prob = plogis(eta_bin))
g_bin$prop <- g_bin$success / g_bin$n
fit_bin <- rsm_glm_fit(g_bin, "prop", c("Dose", "Time"),
                       family = binomial(), weights = g_bin$n)
summary(fit_bin$model)$coefficients
```

```{r glm-binomial-fig, fig.cap="Binomial GLM: observed proportions vs fitted probabilities.", fig.alt="Scatterplot of proportions vs fitted."}
plot(fitted(fit_bin$model), g_bin$prop,
     pch = 19, col = "#009E73",
     xlab = "Fitted probability", ylab = "Observed proportion",
     main = "Binomial GLM: Observed vs Fitted")
abline(0, 1, lty = 2, col = "grey50")
```

## III.3 Gamma: positive continuous

```{r glm-gamma}
set.seed(20260815)
g_gam <- expand.grid(
  x1 = seq(-1, 1, length.out = 7),
  x2 = seq(-1, 1, length.out = 7)
)
eta_gam <- with(g_gam, 1.8 + 0.25 * x1 - 0.20 * x2 -
  0.30 * x1^2 - 0.25 * x2^2)
g_gam$y <- rgamma(nrow(g_gam), shape = 10, scale = exp(eta_gam) / 10)
fit_gamma <- rsm_glm_fit(g_gam, "y", c("x1", "x2"),
                         family = Gamma(link = "log"))
summary(fit_gamma$model)$coefficients
```

```{r glm-gamma-fig, fig.cap="Gamma GLM: observed vs fitted.", fig.alt="Scatterplot of observed vs fitted for Gamma."}
plot(fitted(fit_gamma$model), g_gam$y,
     pch = 19, col = "#56B4E9",
     xlab = "Fitted values", ylab = "Observed",
     main = "Gamma GLM: Observed vs Fitted")
abline(0, 1, lty = 2, col = "grey50")
```


# Part IV: Nonlinear {#sec-nonlinear}

## IV.1 GA initialization

```{r nl-ga}
set.seed(20260816)
d_nl <- expand.grid(
  N = seq(10, 160, length.out = 7),
  P = seq(5, 90, length.out = 7)
)
d_nl$Yield <- with(d_nl,
  11 * (1 - exp(-0.025 * N)) * (1 - exp(-0.040 * P)) +
  rnorm(nrow(d_nl), 0, 0.08)
)
fit_ga <- rsm_nonlinear_fit(
  d_nl, "Yield", c("N", "P"),
  model = "mitscherlich2_product",
  start = "ga",
  start_control = list(popSize = 50, maxiter = 150, run = 50),
  engine = "nlsLM"
)
fit_ga
```

```{r nl-ga-fig, fig.cap="Observed vs fitted for Mitscherlich model with GA initialization.", fig.alt="Scatterplot of observed vs fitted."}
plot(fitted(fit_ga$model), d_nl$Yield,
     pch = 19, col = "#E69F00",
     xlab = "Fitted Yield", ylab = "Observed Yield",
     main = "Mitscherlich (GA start): Observed vs Fitted")
abline(0, 1, lty = 2, col = "grey50")
```

## IV.2 Diagnostics

```{r nl-diag}
rsm_nonlinear_diagnostics(fit_ga)
```

## IV.3 Michaelis-Menten

```{r nl-mm}
d_mm <- expand.grid(
  N = seq(10, 150, length.out = 6),
  P = seq(5, 80, length.out = 6)
)
d_mm$Yield <- with(d_mm, 10 * N / (30 + N) * P / (20 + P))
fit_mm <- rsm_nonlinear_fit(
  d_mm, "Yield", c("N", "P"),
  model = "michaelis_menten2",
  start = c(A = 10, K1 = 30, K2 = 20),
  engine = "nls"
)
rsm_nonlinear_diagnostics(fit_mm)
```

```{r nl-mm-fig, fig.cap="Michaelis-Menten: observed vs fitted.", fig.alt="Scatterplot for Michaelis-Menten."}
plot(fitted(fit_mm$model), d_mm$Yield,
     pch = 19, col = "#CC79A7",
     xlab = "Fitted Yield", ylab = "Observed Yield",
     main = "Michaelis-Menten: Observed vs Fitted")
abline(0, 1, lty = 2, col = "grey50")
```

## IV.4 Optimization

```{r nl-opt}
opt_nl <- rsm_optimize(fit_ga, goal = "max", method = "L-BFGS-B")
cat("Optimal N:", round(opt_nl$solution["N"], 1), "kg/ha\n")
cat("Optimal P:", round(opt_nl$solution["P"], 1), "kg/ha\n")
cat("Predicted Yield:", round(opt_nl$predicted, 2), "Mg/ha\n")
```


# Part V: Tier 3 Surrogates {#sec-tier3}

## V.1 Gaussian Process

```{r tier3-gp}
set.seed(1301)
d_gp <- data.frame(x1 = runif(50, -2, 2), x2 = runif(50, -2, 2))
d_gp$y <- with(d_gp, sin(x1) * cos(x2) + 0.3 * rnorm(nrow(d_gp)))
gp <- rsm_surrogate(d_gp, "y", c("x1", "x2"), method = "gp")
```

```{r tier3-gp-fig, fig.cap="GP surrogate: observed vs predicted.", fig.alt="Scatterplot for GP."}
pr_gp <- predict(gp, d_gp, se.fit = TRUE)
plot(pr_gp$fit, d_gp$y,
     pch = 19, col = "#0072B2",
     xlab = "GP Predicted", ylab = "Observed",
     main = "Gaussian Process: Observed vs Predicted")
abline(0, 1, lty = 2, col = "grey50")
```

## V.2 GAM

```{r tier3-gam}
gam_fit <- rsm_surrogate(d_gp, "y", c("x1", "x2"), method = "gam")
```

```{r tier3-gam-fig, fig.cap="GAM surrogate: observed vs predicted.", fig.alt="Scatterplot for GAM."}
pr_gam <- predict(gam_fit, d_gp)
plot(pr_gam, d_gp$y,
     pch = 19, col = "#009E73",
     xlab = "GAM Predicted", ylab = "Observed",
     main = "GAM: Observed vs Predicted")
abline(0, 1, lty = 2, col = "grey50")
```

## V.3 Random Forest

```{r tier3-rf}
rf_fit <- rsm_surrogate(d_gp, "y", c("x1", "x2"), method = "rf")
```

```{r tier3-rf-fig, fig.cap="Random Forest: observed vs predicted.", fig.alt="Scatterplot for RF."}
pr_rf <- predict(rf_fit, d_gp)
plot(pr_rf, d_gp$y,
     pch = 19, col = "#D55E00",
     xlab = "RF Predicted", ylab = "Observed",
     main = "Random Forest: Observed vs Predicted")
abline(0, 1, lty = 2, col = "grey50")
```

## V.4 Neural Network

```{r tier3-nn}
nn_fit <- rsm_surrogate(d_gp, "y", c("x1", "x2"), method = "nn", size = 3)
```

```{r tier3-nn-fig, fig.cap="Neural Network: observed vs predicted.", fig.alt="Scatterplot for NN."}
pr_nn <- predict(nn_fit, d_gp)
plot(pr_nn, d_gp$y,
     pch = 19, col = "#56B4E9",
     xlab = "NN Predicted", ylab = "Observed",
     main = "Neural Network: Observed vs Predicted")
abline(0, 1, lty = 2, col = "grey50")
```

## V.5 Bayesian Optimization

```{r tier3-bayes}
next_point <- rsm_bayes_opt(d_gp, "y", c("x1", "x2"), n_candidates = 1000)
cat("Recommended next point:\n")
print(next_point$next_point)
cat("Expected improvement:", round(next_point$expected_improvement, 4), "\n")
```

## V.6 Optimization of surrogates

```{r tier3-opt}
opt_gp <- rsm_surrogate_optimize(gp, goal = "max", method = "L-BFGS-B")
cat("GP optimum:", round(opt_gp$solution, 3), "\n")
cat("GP predicted:", round(opt_gp$predicted, 3), "\n")
```


# Part VI: Graphics {#sec-graphics}

## VI.1 Perturbation plot

```{r fig-perturbation, fig.cap="Perturbation plot: sensitivity of yield to each factor.", fig.alt="Perturbation plot."}
rsm_perturbation(fit)
```

## VI.2 FDS plot

```{r fig-fds, fig.cap="Fraction of Design Space.", fig.alt="FDS curve."}
rsm_fds(fit)
```

## VI.3 VDG

```{r fig-vdg, fig.cap="Variance Dispersion Graph.", fig.alt="VDG plot."}
rsm_vdg(fit)
```


# References {.unnumbered}

- Box, G. E. P. & Draper, N. R. (2007). *Response Surfaces*. Wiley.
- Myers, R. H. et al. (2016). *Response Surface Methodology*. Wiley.
- Rasmussen, C. E. & Williams, C. K. I. (2006). *Gaussian Processes for Machine Learning*. MIT Press.

```{r session-info}
sessionInfo()
```
